---
title: "Introduction to Biostatistics"
format:
  # html: 
  #   toc: true
  #   toc-location: left
  #   slide-level: 3
  #   number-sections: true
  #   toc-depth: 2
  #   self-contained: true
  #   #output-file: "Mult22-1-handout"
  # pdf: 
  #   toc: true
  #   number-sections: true
  #   #slide-level: 3
  #   toc-depth: 2
  #   shift-heading-level-by: 0
  #   number-offset: [2]
  revealjs:
    toc: false
    number-sections: true
    #slide-level: 3
    scrollable: true
    embed-resources: true
    slide-number: true
    width: 1200
knitr: 
  opts_chunk:
    echo: true
    message: false
    warning: false
---

```{r, include = FALSE, echo = FALSE, wanring = FALSE, message= FALSE}
library(tidyverse)
theme_set(theme_classic(base_size = 16))
set.seed(12345)
options(scipen= 100)
```

# Learning objectives

## Learning objectives

* *Interpret* tests comparing **two unrelated samples**
* *Summarize data* using **contingency tables**
* *Describe* different study designs for **contingency tables** 

# Two independent samples

## Independent samples tests

* Compare **means** (or *medians*) from two *unrelated* samples
* Do the two samples come from populations with the same mean?
  * Is the difference between the two population means 0?

## Independent samples tests

* Parametric tests
  * Independent samples $z$-test
  * Independent samples $t$-test (also Welch's $t$-test)
* Non-parametric tests
  * Median test
  * Wilcoxon-Mann-Whitney U test
  * Chi-square test and Fisher's exact test (*next week*)

## Independence

* Statistical independence
  * Two events are **independent** if the occurrence of one event doesn't impact the occurrence of another event
  * Knowing about one event tells you nothing about the other event
* Independent samples
  * Individuals in one sample are **not related** to individuals in the other sample
  * The two samples are made up of **different** individuals

## $z$-test: Assumptions

* Data are **continuous** (i.e., ratio or interval)
* Data are **randomly sampled** from the population(s)
* Data are **independent** (*within* and *across* groups)
* Data are **approximately normally distributed OR** sample size is large enough for **normally distributed sampling distribution** (*central limit theorem*)
* Population variance (or SD) is **known** and **same in both groups**
  * "Large sample": population variance estimated by sample variance

## $z$-test: Hypotheses

* Directional (one-tailed) tests
  * $H_0$: $\mu_1 \le \mu_2$ or $\mu_1 - \mu_2 \le 0$
    * $H_1$: $\mu_1 > \mu_2$ or $\mu_1 - \mu_2 > 0$
  * $H_0$: $\mu_1 \ge \mu_2$ or $\mu_1 - \mu_2 \ge 0$
    * $H_1$: $\mu_1 < \mu_2$ or $\mu_1 - \mu_2 < 0$
* Non-directional (two-tailed) tests
  * $H_0$: $\mu_1 = \mu_2$ or $\mu_1 - \mu_2 = 0$
    * $H_1$: $\mu_1 \ne \mu_2$ or $\mu_1 - \mu_2 \ne 0$

## $z$-test: Test statistic

$$z = \frac{(\bar{X_1} - \bar{X_2}) - (\mu_1 - \mu_2)}{\sqrt{\frac{\sigma^2_1}{n_1} + \frac{\sigma^2_2}{n_2}}}$$ 

* $\mu_1 - \mu_2 = 0$ (according to $H_0$)
* $\sigma^2_1$ and $\sigma^2_2$ are the variances of the two groups
* $n_1$ and $n_2$ are the sample sizes of the two groups

## $z$-test: Example 1

* Is resting pulse rate the same for smokers and non-smokers?
  * One dataset / variable (column) for each group

:::: {.columns}

::: {.column width="50%"}

```{r}
#| code-fold: true
library(Stat2Data)
data(Pulse)
#head(Pulse)
library(tidyverse)
Pulse_smoke <- Pulse %>% filter(Smoke == 1)
Pulse_nosmoke <- Pulse %>% filter(Smoke == 0)
head(Pulse_smoke)
```

:::

::: {.column width="50%"}

```{r}
#| code-fold: true
head(Pulse_nosmoke)
```

:::

::::

## $z$-test: Example 2

* Is resting pulse rate the same for smokers and non-smokers?
  * Check out the data

Group       | n     | Mean  | SD    
------------|-------|-------|-------
Non-smokers | `r nrow(Pulse_nosmoke)` | `r round(mean(Pulse_nosmoke$Rest), 3)` | `r round(sd(Pulse_nosmoke$Rest), 3)`
Smokers     | `r nrow(Pulse_smoke)` | `r round(mean(Pulse_smoke$Rest), 3)` | `r round(sd(Pulse_smoke$Rest), 3)`

## $z$-test: Example 3

* Is resting pulse rate the same for smokers and non-smokers?
  * Check out the data
  
```{r}
#| code-fold: true
ggplot(data = Pulse_nosmoke, 
       aes(x = Rest)) +
  geom_histogram(fill = "red", 
                 alpha = 0.5, 
                 bins = 30) +
  geom_histogram(data = Pulse_smoke, 
                 aes(x = Rest), 
                 fill = "black", 
                 bins = 30)
```

## $z$-test: Example 4

* Is resting pulse rate the same for smokers and non-smokers?
  * `z.test()` function from **BSDA** package

```{r}
library(BSDA)
ztest <- z.test(x = Pulse_smoke$Rest, 
                y = Pulse_nosmoke$Rest, 
                sigma.x = sd(Pulse_smoke$Rest), 
                sigma.y = sd(Pulse_nosmoke$Rest), 
                alternative = "two.sided")
ztest
```

## $z$-test: Report results

* Reject $H_0$: $\mu_1 = \mu_2$ or $\mu_1 - \mu_2 = 0$
  * $p$-value < .05 
  * 95% confidence interval doesn't contain 0 (value from $H_0$)
  * These two samples came from *different* populations
* Using a two-sample $z$-test, we rejected the null hypothesis that the means are equal, $z$ = 2.44, $p$ = .01
  * Smokers and non-smokers have *different* resting pulse rates

## $t$-test: Assumptions

* Data are **continuous** (i.e., ratio or interval)
* Data are **randomly sampled** from the population
* Data are **independent** (*within* and *across* groups)
* Data are **approximately normally distributed OR** sample size is large enough for **normally distributed sampling distribution** (*central limit theorem*)
* Population variance (or SD) is **unknown** and **same in both groups**

## $t$-test: Hypotheses

* Directional (one-tailed) tests
  * $H_0$: $\mu_1 \le \mu_2$ or $\mu_1 - \mu_2 \le 0$
    * $H_1$: $\mu_1 > \mu_2$ or $\mu_1 - \mu_2 > 0$
  * $H_0$: $\mu_1 \ge \mu_2$ or $\mu_1 - \mu_2 \ge 0$
    * $H_1$: $\mu_1 < \mu_2$ or $\mu_1 - \mu_2 < 0$
* Non-directional (two-tailed) tests
  * $H_0$: $\mu_1 = \mu_2$ or $\mu_1 - \mu_2 = 0$
    * $H_1$: $\mu_1 \ne \mu_2$ or $\mu_1 - \mu_2 \ne 0$

## $t$-test: Test statistic

$$t = \frac{(\bar{X_1} - \bar{X_2}) - (\mu_1 - \mu_2)}{\sqrt{\frac{s^2_p}{n_1} + \frac{s^2_p}{n_2}}}$$ 

* where $s^2_p = \frac{(n_1 -1)s^2_1 + (n_2 -1)s^2_2}{n_1 + n_2 -2}$
  * Assumes equal variances and pools (combines)
* with degrees of freedom = $n_1 + n_2 - 2$

## $t$-test: Example 1

* Is resting pulse rate the same for smokers and non-smokers?
  * `t.test()` function from **stats** package

```{r}
ttest <- t.test(x = Pulse_smoke$Rest, 
                y = Pulse_nosmoke$Rest, 
                alternative = "two.sided",
                var.equal = TRUE)
ttest
```

## $t$-test: Report results

* Reject $H_0$: $\mu_1 = \mu_2$ or $\mu_1 - \mu_2 = 0$
  * $p$-value < .05 
  * 95% confidence interval doesn't contain 0 (value from $H_0$)
  * These two samples came from *different* populations
* Using a independent samples $t$-test, we rejected the null hypothesis that the means are equal, $t(230)$ = 2.43, $p$ = .02
  * Smokers and non-smokers have *different* resting pulse rates

## Welch's $t$-test: Unequal variances

$$t = \frac{(\bar{X_1} - \bar{X_2}) - (\mu_1 - \mu_2)}{\sqrt{\frac{s^2_1}{n_1} + \frac{s^2_2}{n_2}}}$$ 

* with degrees of freedom = **something much more complicated** 
  * May be fractional

## Welch's $t$-test: Example

```{r}
welch <- t.test(x = Pulse_smoke$Rest, 
                y = Pulse_nosmoke$Rest, 
                alternative = "two.sided",
                var.equal = FALSE)
welch
```

## Welch's $t$-test: Report results

* Reject $H_0$: $\mu_1 = \mu_2$ or $\mu_1 - \mu_2 = 0$
  * $p$-value < .05 
  * 95% confidence interval doesn't contain 0 (value from $H_0$)
  * These two samples came from *different* populations
* Using a Welch's $t$-test, we rejected the null hypothesis that the means are equal, $t(31.721)$ = 2.44, $p$ = .02
  * Smokers and non-smokers have *different* resting pulse rates

## Very similar results

Test | Diff | Statistic | df | $p$ | CI
------|------|------|------|------|------
$z$   | `r round(mean(Pulse_smoke$Rest) - mean(Pulse_nosmoke$Rest), 3)` | `r round(ztest$statistic, 3)` | NA | `r round(ztest$p.value, 3)` | [`r round(ztest$conf.int, 3)`]
$t$   | `r round(mean(Pulse_smoke$Rest) - mean(Pulse_nosmoke$Rest), 3)` | `r round(ttest$statistic, 3)` | `r round(ttest$parameter, 3)` | `r round(ttest$p.value, 3)` | [`r round(ttest$conf.int, 3)`]
Welch | `r round(mean(Pulse_smoke$Rest) - mean(Pulse_nosmoke$Rest), 3)` | `r round(welch$statistic, 3)` | `r round(welch$parameter, 3)` | `r round(welch$p.value, 3)` | [`r round(welch$conf.int, 3)`]

## Non-parametric tests

* Non-parametric versions of $z$ or $t$-test for 2 independent samples
  * Median test
  * Wilcoxon-Mann-Whitney U test (or Mann-Whitney or Mann-Whitney-Wilcoxon)
* Differences in **median** (technically, "location")
* Mann-Whitney is a somewhat better test

## Non-parametric tests

:::: {.columns}

::: {.column width="50%"}

* Median test

```{r}
#| code-fold: true
library(coin)
median_test(Rest ~ as.factor(Smoke), data = Pulse)
```

:::

::: {.column width="50%"}

* Wilcoxon-Mann-Whitney test

```{r}
#| code-fold: true
wilcox.test(Rest ~ as.factor(Smoke), data = Pulse)
```

:::

::::

# Contingency tables

## Contingency tables

* Cross-tabs, summary tables, 2x2 table
  * Relationship between two (or more) categorical variables
  * Each cell is a **frequency** for that combination
* `Sex` and `Smoke` from the `Pulse` dataset

```{r}
#| code-fold: true
smoke_sex <- table(Pulse$Sex, Pulse$Smoke)
colnames(smoke_sex) <- c("Non-smoker", "Smoker")
rownames(smoke_sex) <- c("Male", "Female")
smoke_sex_margins <- addmargins(smoke_sex)
smoke_sex_prop_margins <- addmargins(prop.table(smoke_sex))
smoke_sex
```

## Notation for frequencies

&nbsp;   | $J = 1$    | $J = 2$    | &nbsp;
---------|:----------:|:----------:|:-----------:
$I = 1$  | $n_{11}$   | $n_{12}$   | $\color{white}{n_{1+}}$
$I = 2$  | $n_{21}$   | $n_{22}$   | &nbsp;
&nbsp;   | &nbsp;     | &nbsp;     | &nbsp;

## Notation for frequencies

&nbsp;   | $J = 1$    | $J = 2$    | &nbsp;
---------|:----------:|:----------:|:-----------:
$I = 1$  | $n_{11}$   | $n_{12}$   | $n_{1+}$
$I = 2$  | $n_{21}$   | $n_{22}$   | $n_{2+}$
&nbsp;   | $n_{+1}$   | $n_{+2}$   | $n$

* $n_{11}$, $n_{12}$, $n_{21}$, $n_{22}$ are **joint** frequencies
* $n_{1+}$, $n_{2+}$, $n_{+1}$, $n_{+2}$ are **marginal** frequencies

## Notation for probabilities

&nbsp;   | $J = 1$    | $J = 2$    | &nbsp;
---------|:----------:|:----------:|:-----------:
$I = 1$  | $p_{11}$   | $p_{12}$   | $\color{white}{p_{1+}}$
$I = 2$  | $p_{21}$   | $p_{22}$   | &nbsp;
&nbsp;   | &nbsp;     | &nbsp;     | &nbsp;

## Notation for probabilities

&nbsp;   | $J = 1$    | $J = 2$    | &nbsp;
---------|:----------:|:----------:|:-----------:
$I = 1$  | $p_{11}$   | $p_{12}$   | $p_{1+}$
$I = 2$  | $p_{21}$   | $p_{22}$   | $p_{2+}$
&nbsp;   | $p_{+1}$   | $p_{+2}$   | $1$

* $p_{11}$, $p_{12}$, $p_{21}$, $p_{22}$ are **joint** probabilities = $\frac{n_{ij}}{n}$
* $p_{1+}$, $p_{2+}$, $p_{+1}$, $p_{+2}$ are **marginal** probabilities = $\frac{n_{i+}}{n}$ or $\frac{n_{+j}}{n}$

## Marginal probability

* **Marginal probability**: Probability of $X$ or $Y$, collapsing over the other
  * What is the distribution of $X$, ignoring $Y$?
  * What is the distribution of $Y$, ignoring $X$?
 
## Marginal probability

* Start with frequencies

&nbsp;          | $Y$ = 1: No smoke   | $Y$ = 2: Smoke    | &nbsp;
------------------|:------------------:|:--------------:|:--------------:
$X = 1$: Male  | $n_{11} = \color{red}{`r smoke_sex_margins[1,1]`}$ | $n_{12} = \color{red}{`r smoke_sex_margins[1,2]`}$ | $n_{1+} = \color{blue}{`r smoke_sex_margins[1,3]`}$
$X = 2$: Female   | $n_{21} = \color{red}{`r smoke_sex_margins[2,1]`}$ | $n_{22} = \color{red}{`r smoke_sex_margins[2,2]`}$ | $n_{2+} = \color{blue}{`r smoke_sex_margins[2,3]`}$
&nbsp;          | $n_{+1} = \color{blue}{`r smoke_sex_margins[3,1]`}$ | $n_{+2} = \color{blue}{`r smoke_sex_margins[3,2]`}$ | $n = \color{blue}{`r smoke_sex_margins[3,3]`}$

## Marginal probability

* Divide each value by $n$: The total sample size    

&nbsp;          | $Y$ = 1: No smoke   | $Y$ = 2: Smoke    | &nbsp;
------------------|:------------------:|:--------------:|:--------------:
$X = 1$: Male  | $p_{11} = \color{red}{`r round(smoke_sex_prop_margins[1,1], 2)`}$ | $p_{12} = \color{red}{`r round(smoke_sex_prop_margins[1,2], 2)`}$ | $p_{1+} = \color{blue}{`r round(smoke_sex_prop_margins[1,3], 2)`}$
$X = 2$: Female   | $p_{21} = \color{red}{`r round(smoke_sex_prop_margins[2,1], 2)`}$ | $p_{22} = \color{red}{`r round(smoke_sex_prop_margins[2,2], 2)`}$ | $p_{2+} = \color{blue}{`r round(smoke_sex_prop_margins[2,3], 2)`}$
&nbsp;          | $p_{+1} = \color{blue}{`r round(smoke_sex_prop_margins[3,1], 2)`}$ | $p_{+2} = \color{blue}{`r round(smoke_sex_prop_margins[3,2], 2)`}$ | $p = \color{blue}{`r round(smoke_sex_prop_margins[3,3], 2)`}$

* $\color{red}{Joint}$ probabilities sum to 1
* $\color{blue}{Marginal}$ probabilities for *rows* sum to 1
* $\color{blue}{Marginal}$ probabilities for *columns* sum to 1

## Conditional probability

* Often, $X$ is an explanatory variable, $Y$ is an outcome variable
  * **But it doesn't need to be**

* **Conditional probability**: Probability of $Y$ **at a given value of $X$**
  * When $X = 1$, what is the distribution of $Y$?
  * When $X = 2$, what is the distribution of $Y$?

* Conditional probability is the $\color{red}{joint~value}$ divided by the $\color{blue}{marginal~value}$ **for that value of $X$**
  * It is **conditional** on that value of $X$

## Conditional probability

&nbsp;          | $Y$ = 1: No smoke   | $Y$ = 2: Smoke    | &nbsp;
------------------|:------------------:|:--------------:|:--------------:
$X = 1$: Male  | $n_{11} = \color{red}{`r smoke_sex_margins[1,1]`}$ | $n_{12} = \color{red}{`r smoke_sex_margins[1,2]`}$ | $n_{1+} = \color{blue}{`r smoke_sex_margins[1,3]`}$
$X = 2$: Female   | $n_{21} = \color{red}{`r smoke_sex_margins[2,1]`}$ | $n_{22} = \color{red}{`r smoke_sex_margins[2,2]`}$ | $n_{2+} = \color{blue}{`r smoke_sex_margins[2,3]`}$
&nbsp;          | $n_{+1} = \color{blue}{`r smoke_sex_margins[3,1]`}$ | $n_{+2} = \color{blue}{`r smoke_sex_margins[3,2]`}$ | $n = \color{blue}{`r smoke_sex_margins[3,3]`}$

## Conditional probability

&nbsp;          | $Y$ = 1: No smoke   | $Y$ = 2: Smoke    | &nbsp;
------------------|:------------------:|:--------------:|:--------------:
$X = 1$: Male  | $n_{11} = \color{red}{`r smoke_sex_margins[1,1]`}$ | $n_{12} = \color{red}{`r smoke_sex_margins[1,2]`}$ | $n_{1+} = \color{blue}{`r smoke_sex_margins[1,3]`}$
$X = 2$: Female   | $n_{21} = \color{red}{`r smoke_sex_margins[2,1]`}$ | $n_{22} = \color{red}{`r smoke_sex_margins[2,2]`}$ | $n_{2+} = \color{blue}{`r smoke_sex_margins[2,3]`}$
&nbsp;          | $n_{+1} = \color{blue}{`r smoke_sex_margins[3,1]`}$ | $n_{+2} = \color{blue}{`r smoke_sex_margins[3,2]`}$ | $n = \color{blue}{`r smoke_sex_margins[3,3]`}$

:::: {.columns}

::: {.column width="50%"}
* When $X = 1$ (male):
  * $P(no\ smoke) = \frac{\color{red}{`r smoke_sex_margins[1,1]`}}{\color{blue}{`r smoke_sex_margins[1,3]`}} = `r round(smoke_sex_margins[1,1]/smoke_sex_margins[1,3], 3)`$
  * $P(smoke) = \frac{\color{red}{`r smoke_sex_margins[1,2]`}}{\color{blue}{`r smoke_sex_margins[1,3]`}} = `r round(smoke_sex_margins[1,2]/smoke_sex_margins[1,3], 3)`$
:::

::: {.column width="50%"}
* When $X = 2$ (female):
  * $P(no\ smoke) = \frac{\color{red}{`r smoke_sex_margins[2,1]`}}{\color{blue}{`r smoke_sex_margins[2,3]`}} = `r round(smoke_sex_margins[2,1]/smoke_sex_margins[2,3], 3)`$
  * $P(smoke) = \frac{\color{red}{`r smoke_sex_margins[2,2]`}}{\color{blue}{`r smoke_sex_margins[2,3]`}} = `r round(smoke_sex_margins[2,2]/smoke_sex_margins[2,3], 3)`$
:::

::::

## `Sex` and `Smoke`: Frequencies

```{r}
#| code-fold: true
smoke_sex
```

## `Sex` and `Smoke`: And margins

```{r}
#| code-fold: true
smoke_sex_margins
```

## `Sex` and `Smoke`: Marginal prob

```{r}
#| code-fold: true
smoke_sex_prop_margins
```

## `Sex` and `Smoke`: Conditional prob

```{r}
#| code-fold: true
prop.table(smoke_sex, margin = 1)
```

* Conditional on `Sex`

# Study design

## Fixed vs random

* The *marginal frequencies* of a contingency table can be either **fixed** or **random**
  * **Fixed**: Chosen by the researcher
    * e.g., Collect data on 50 men and 50 women
  * **Random**: Vary depending on the sample
    * e.g., Collect data on gender in the sample

## Why do we care?

* **Probability** = random divided by fixed
  * If there are no fixed marginals, **we can’t calculate a probability**
* **Ratio** = random divided by random
  * We can always calculate ratios (e.g., odds ratios)
* What is “fixed” is what you can “condition on”
  * What are probabilities "out of"?

## Why do we care?

* Design of the contingency table determines 
  * What is **conditioned on** 
  * What are **probabilities** (and what are just *ratios*)
  * How you can talk about the **relationship** in the table
  * What **statistical tests** you can perform

## Types of study designs

* Three study designs with different fixed and random marginals
  * Cross-sectional (or multinomial)
  * Retrospective
  * Prospective (or product binomial)

## Overall study design

&nbsp;  | Heart attack | No heart attack | $\color{white}{White text}$
--------|:------------:|:---------------:|:--------:
Placebo | &nbsp;       | &nbsp;          | &nbsp;
Aspirin | &nbsp;       | &nbsp;          | &nbsp;
&nbsp;  | &nbsp;       | &nbsp;          | &nbsp;

* Relationship between aspirin use (vs placebo) and heart attack
  * $X$: Aspirin vs placebo
  * $Y$: Heart attack vs no heart attack

## Cross-sectional

&nbsp;  | Heart attack | No heart attack | $\color{white}{White text}$
--------|:------------:|:---------------:|:--------:
Placebo | &nbsp;       | &nbsp;          | Random
Aspirin | &nbsp;       | &nbsp;          | Random
&nbsp;  | Random       | Random          | **Fixed**

* Collect data from $n$ people
  * Measure aspirin vs placebo, heart attack vs not

## Retrospective

&nbsp;  | Heart attack | No heart attack | $\color{white}{White text}$
--------|:------------:|:---------------:|:--------:
Placebo | &nbsp;       | &nbsp;          | Random
Aspirin | &nbsp;       | &nbsp;          | Random
&nbsp;  | **Fixed**    | **Fixed**       | &nbsp;

* Collect data from specific numbers of heart attack and non patients
  * Measure whether they took aspirin

## Prospective

&nbsp;  | Heart attack | No heart attack | $\color{white}{White text}$
--------|:------------:|:---------------:|:--------:
Placebo | &nbsp;       | &nbsp;          | **Fixed**
Aspirin | &nbsp;       | &nbsp;          | **Fixed**
&nbsp;  | Random       | Random          | &nbsp;

* Collect data from specific number of aspirin and placebo people
  * Measure whether they have a heart attack

## `Smoke` and `Sex` example

* What type of study design was this?
  * Multinomial: Total $n$ is fixed
  * Retrospective: `Smoke` is fixed
  * Prospective: `Sex` is fixed
  
```{r}
#| code-fold: true
smoke_sex_margins
```

## Measures of relationship

* Any design
  * Odds ratio
  * Test of independence
* Prospective design only
  * Difference in proportion
  * Relative risk
  * Odds ratio
  * Test of independence

## Difference in proportion

* If we assume it is a **prospective** design re: `Sex`

:::: {.columns}

::: {.column width="50%"}

* Male
  * $P(smoke) = \frac{\color{red}{`r smoke_sex_margins[1,2]`}}{\color{blue}{`r smoke_sex_margins[1,3]`}} = `r round(smoke_sex_margins[1,2]/smoke_sex_margins[1,3], 3)`$

:::

::: {.column width="50%"}

* Female
  * $P(smoke) = \frac{\color{red}{`r smoke_sex_margins[2,2]`}}{\color{blue}{`r smoke_sex_margins[2,3]`}} = `r round(smoke_sex_margins[2,2]/smoke_sex_margins[2,3], 3)`$

:::

::::

* Difference in proportion = $`r round(smoke_sex_margins[1,2]/smoke_sex_margins[1,3], 3)` - `r round(smoke_sex_margins[2,2]/smoke_sex_margins[2,3], 3)` = `r round(smoke_sex_margins[1,2]/smoke_sex_margins[1,3] - smoke_sex_margins[2,2]/smoke_sex_margins[2,3], 3)`$
  * A man is `r round(100*(smoke_sex_margins[1,2]/smoke_sex_margins[1,3] - smoke_sex_margins[2,2]/smoke_sex_margins[2,3]), 1)` percentage points more likely to be a smoker than a woman

## Relative risk

* If we assume it is a **prospective** design re: `Sex`

:::: {.columns}

::: {.column width="50%"}

* Male
  * $P(smoke) = \frac{\color{red}{`r smoke_sex_margins[1,2]`}}{\color{blue}{`r smoke_sex_margins[1,3]`}} = `r round(smoke_sex_margins[1,2]/smoke_sex_margins[1,3], 3)`$

:::

::: {.column width="50%"}

* Female
  * $P(smoke) = \frac{\color{red}{`r smoke_sex_margins[2,2]`}}{\color{blue}{`r smoke_sex_margins[2,3]`}} = `r round(smoke_sex_margins[2,2]/smoke_sex_margins[2,3], 3)`$

:::

::::

* Relative risk = $\frac{`r round(smoke_sex_margins[1,2]/smoke_sex_margins[1,3], 3)`}{`r round(smoke_sex_margins[2,2]/smoke_sex_margins[2,3], 3)`} = `r round((smoke_sex_margins[1,2]/smoke_sex_margins[1,3]) / (smoke_sex_margins[2,2]/smoke_sex_margins[2,3]), 3)`$
  * A man is `r round((smoke_sex_margins[1,2]/smoke_sex_margins[1,3]) / (smoke_sex_margins[2,2]/smoke_sex_margins[2,3]), 3)` times more likely to be a smoker than a woman

## Odds ratio

$$odds(smoke) = P(smoke)/P(no\ smoke)$$

:::: {.columns}

::: {.column width="50%"}

* Male
  * $P(smoke) = \frac{\color{red}{`r smoke_sex_margins[1,2]`}}{\color{blue}{`r smoke_sex_margins[1,3]`}} = `r round(smoke_sex_margins[1,2]/smoke_sex_margins[1,3], 3)`$
  * $odds = \frac{`r round(smoke_sex_margins[1,2]/smoke_sex_margins[1,3], 3)`}{`r round(1 - smoke_sex_margins[1,2]/smoke_sex_margins[1,3], 3)`} = `r round(smoke_sex_margins[1,2]/smoke_sex_margins[1,3]/(1 - smoke_sex_margins[1,2]/smoke_sex_margins[1,3]), 3)`$

:::

::: {.column width="50%"}

* Female
  * $P(smoke) = \frac{\color{red}{`r smoke_sex_margins[2,2]`}}{\color{blue}{`r smoke_sex_margins[2,3]`}} = `r round(smoke_sex_margins[2,2]/smoke_sex_margins[2,3], 3)`$
  * $odds = \frac{`r round(smoke_sex_margins[2,2]/smoke_sex_margins[2,3], 3)`}{`r round(1 - smoke_sex_margins[2,2]/smoke_sex_margins[2,3], 3)`} = `r round(smoke_sex_margins[2,2]/smoke_sex_margins[2,3] / (1 - smoke_sex_margins[2,2]/smoke_sex_margins[2,3]), 3)`$

:::

::::

* $odds\ ratio = \frac{`r round(smoke_sex_margins[2,2]/smoke_sex_margins[2,3] / (1 - smoke_sex_margins[2,2]/smoke_sex_margins[2,3]), 3)`}{`r round(smoke_sex_margins[1,2]/smoke_sex_margins[1,3]/(1 - smoke_sex_margins[1,2]/smoke_sex_margins[1,3]), 3)`} = 
`r round(((smoke_sex_margins[2,2]/smoke_sex_margins[2,3]) / (1 - smoke_sex_margins[2,2]/smoke_sex_margins[2,3]))/(smoke_sex_margins[1,2]/smoke_sex_margins[1,3]/(1 - smoke_sex_margins[1,2]/smoke_sex_margins[1,3])), 3)`$
  * The **odds** of a woman smoking is `r round(((smoke_sex_margins[2,2]/smoke_sex_margins[2,3]) / (1 - smoke_sex_margins[2,2]/smoke_sex_margins[2,3]))/(smoke_sex_margins[1,2]/smoke_sex_margins[1,3]/(1 - smoke_sex_margins[1,2]/smoke_sex_margins[1,3])), 3)` times the **odds** of a man smoking

## Independence

* Do the **observed** frequencies match what we'd *expect* if the variables were **unrelated** or **independent**?

:::: {.columns}

::: {.column width="50%"}

* Observed

```{r}
#| code-fold: true
smoke_sex
```

:::

::: {.column width="50%"}

* Expected (independence)

```{r}
#| code-fold: true
library(epitools)
expected(smoke_sex)
```

:::

::::

# In-class activities

## In-class activities

* Conduct two independent samples tests
  * $z$-test, $t$-test, Welch's $t$-test
* Look at contingency tables and orient to them
* Think about study design

## Next week

* Relate study design to tests for contingency tables
  * Some tests are only available for some designs
* Related-samples tests
