---
title: "Introduction to Biostatistics"
format:
  # html: 
  #   toc: true
  #   toc-location: left
  #   slide-level: 3
  #   number-sections: true
  #   toc-depth: 2
  #   self-contained: true
  #   #output-file: "Mult22-1-handout"
  # pdf: 
  #   toc: true
  #   number-sections: true
  #   #slide-level: 3
  #   toc-depth: 2
  #   shift-heading-level-by: 0
  #   number-offset: [2]
  revealjs:
    toc: false
    number-sections: true
    #slide-level: 3
    scrollable: true
    embed-resources: true
    slide-number: true
    width: 1200
knitr: 
  opts_chunk:
    echo: true
    message: false
    warning: false
---

```{r, include = FALSE, echo = FALSE, wanring = FALSE, message= FALSE}
library(tidyverse)
library(grid)
library(xtable)
library(kableExtra)
theme_set(theme_classic(base_size = 16))
set.seed(12345)
options(scipen= 100)
```

# Learning objectives

## Learning objectives

* *Describe* the logic of ANOVA
* *Conduct* and *interpret* the overall ANOVA test
* *Conduct* post hoc tests

# Extending the $t$-test

## Review: $t$-test

* Compare exactly **2 groups**
* Focus on **difference** between group **means**
* $t$ distribution
* Statistical test: difference / standard error

## ANOVA compares 3+ groups

* Medication: Control, low dose, high dose
* Treatment: Usual care, new treatment 1, new treatment 2
* Modality: In-person, online, hybrid

## 2 group approaches...

```{r}
#| code-fold: true
ggplot(data.frame(x = c(-3, 7)), aes(x)) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1), 
                geom = "polygon", 
                alpha = 0.3, 
                fill = "blue") +
  stat_function(fun = dnorm, args = list(mean = 2, sd = 1), 
                geom = "polygon", 
                alpha = 0.3,
                fill = "red") +
  ylim(0,.5) + 
  scale_x_continuous(breaks = -3:7) +
  labs(x = "X", y = "") +
  annotate("text", x = 0, y = 0.2, label = "Group 1", color = "blue") +
  annotate("text", x = 2, y = 0.2, label = "Group 2", color = "red") +
  annotate("segment", x = 0,, xend = 2, y = .4, yend = .4,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
  annotate("text", x = 1, y = .47, label = "Difference")
```

## ... don't extend well to 3+ groups

```{r}
#| code-fold: true
ggplot(data.frame(x = c(-3, 7)), aes(x)) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1), 
                geom = "polygon", 
                alpha = 0.3, 
                fill = "blue") +
  stat_function(fun = dnorm, args = list(mean = 2, sd = 1), 
                geom = "polygon", 
                alpha = 0.3,
                fill = "red") +
  stat_function(fun = dnorm, args = list(mean = 4, sd = 1), 
                geom = "polygon", 
                alpha = 0.3,
                fill = "forestgreen") +
  ylim(0,.5) + 
  scale_x_continuous(breaks = -3:7) +
  labs(x = "X", y = "") +
  annotate("text", x = 0, y = 0.2, label = "Group 1", color = "blue") +
  annotate("text", x = 2, y = 0.2, label = "Group 2", color = "red") +
  annotate("text", x = 4, y = 0.2, label = "Group 3", color = "forestgreen") +
  annotate("segment", x = 0, xend = 1.95, y = .4, yend = .4,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
  annotate("text", x = 2, y = .47, label = "Differences?") +
  annotate("segment", x = 2.05, xend = 4, y = .4, yend = .4,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
  annotate("segment", x = 0, xend = 4, y = .43, yend = .43,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm")))
```

# Between-subjects ANOVA

## Logic

* Extension of independent samples $t$-test
* **Sources of variability** instead of differences
  * Variability *across groups*
  * Variability across individuals *within a group* (random error)
* Is the *variability across groups* (i.e., mean differences) **large** in the context of the *within-group variability*?
  * Signal to noise ratio

## Between > within

```{r}
#| code-fold: true
ggplot(data.frame(x = c(-3, 7)), aes(x)) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1), 
                geom = "polygon", 
                alpha = 0.3, 
                fill = "blue") +
  stat_function(fun = dnorm, args = list(mean = 4, sd = 1), 
                geom = "polygon", 
                alpha = 0.3,
                fill = "red") +
  ylim(0,.5) + 
  scale_x_continuous(breaks = -3:7) +
  labs(x = "X", y = "") 
```

## Within > between

```{r}
#| code-fold: true
ggplot(data.frame(x = c(-3, 7)), aes(x)) +
  stat_function(fun = dnorm, args = list(mean = 1.5, sd = 2), 
                geom = "polygon", 
                alpha = 0.3, 
                fill = "blue") +
  stat_function(fun = dnorm, args = list(mean = 2.5, sd = 2), 
                geom = "polygon", 
                alpha = 0.3,
                fill = "red") +
  ylim(0,.5) + 
  scale_x_continuous(breaks = -3:7) +
  labs(x = "X", y = "") 
```

## Total variance

```{r}
#| code-fold: true
library(Stat2Data)
data(Amyloid)
#head(Amyloid)
ggplot(data = Amyloid, aes(x = Abeta)) +
  geom_dotplot(method = "histodot") +
  theme(legend.position="none")
```

## Partitioned variance

```{r}
#| code-fold: true
ggplot(data = Amyloid, aes(x = Abeta, fill= Group)) +
  geom_dotplot(method = "histodot", stackgroups = TRUE) +
  theme(legend.position="none")
```

## Partitioned variance

```{r}
#| code-fold: true
means <- Amyloid %>% group_by(Group) %>% summarize(meanAbeta = mean(Abeta))
ggplot(data = Amyloid, aes(x = Abeta, fill= Group)) +
  geom_dotplot(method = "histodot", stackgroups = TRUE) +
  theme(legend.position="none") +
  geom_vline(aes(xintercept = means$meanAbeta[1]), 
             color = "#F8766D",
             linewidth = 1) +
  geom_vline(aes(xintercept = means$meanAbeta[2]), 
             color = "#00BA38",
             linewidth = 1) +
  geom_vline(aes(xintercept = means$meanAbeta[3]), 
             color = "#619CFF",
             linewidth = 1)
```

## Assumptions

* One *categorical* (grouping) variable with 3+ categories
* One *continuous* (i.e., interval or ratio) variable
* Data are *randomly sampled* from the population
* Data are *independent* (within and across groups)
* Data are approximately *normally distributed* OR *sample size is large enough* for normally distributed sampling distribution
  * Residuals from model are **normally distributed**
* Population *variance* is unknown and *same in all groups*

## Hypotheses

* $H_0$: All means are equal
  * $\mu_1 = \mu_2 = \mu_3$
* $H_1$: **At least one mean** is different from the others
  * (**NOT** $\mu_1 \ne \mu_2 \ne \mu_3$)

## ANOVA summary table

Source of variation | Sums of squares (SS) | df | Mean squares (MS) | F
------|------|------|------|------
Between | $SS_{between}$ | $k - 1$ | $MS_{between}$ | $F$
Error | $SS_{error}$ | $N - k$ | $MS_{error}$
Total | $SS_{total}$ | $N - 1$

* $k$ = number of groups
* $N$ = total number of participants
* $n$ = number of participants in a group

## Test statistic 1: Sums of squares

* $SS_{between}$ = variation of $\color{red}{group\ means}$ around $\color{blue}{grand\ mean}$
  * $SS_{between} = \Sigma_{1}^{k} (\color{red}{\bar{X}_j} - \color{blue}{\bar{X}})^2$
* $SS_{error}$ = variation of $individual\ values$ around $\color{red}{group\ means}$
  * $SS_{error} = \Sigma_{1}^{k}\Sigma_{1}^{n} (X - \color{red}{\bar{X}_j})^2$
* $SS_{total}$ = variation of $individual\ values$ around $\color{blue}{grand\ mean}$
  * $SS_{total} = \Sigma_{1}^N (X - \color{blue}{\bar{X}})^2$
* $SS_{total} = SS_{between} + SS_{error}$

## Test statistic 2: Mean squares

Mean squares = Sum of squares divided by degrees of freedom

* $MS_{between}$ = **between group variance**
  * $MS_{between} = \frac{SS_{between}}{k - 1}$
* $MS_{error}$ = **within group variance**
  * $MS_{error} = \frac{SS_{error}}{N-k}$

## Test statistic 3: $F$-statistic

Analysis of "variance" = compare **variances**

* $F$-statistic = **between group variance** / **within group variance**
  * $F = \frac{MS_{between}}{MS_{error}}$
* Compare to critical $F$ from a distribution of $k - 1$ and $N - k$ degrees of freedom
  * If the observed $F$ statistic is larger, reject $H_0$
  * Otherwise, retain $H_0$

## $F$ distribution

* Two different degrees of freedom values
  * First is about the number of **groups**: $k - 1$
  * Second is about the number of **subjects (and groups)**: $N - k$
  
::: {.callout-note}

* Compare to $t$ distribution
  * Just a single df value
  * Always only comparing 2 groups
  * $[t(df)]^2 = F(1, df)$
  
:::

## $F$ distribution

```{r}
#| code-fold: true
ggplot(data.frame(x = c(0, 5)), aes(x)) +
  stat_function(fun = df, 
                args = list(df1 = 2, df2 = 47), 
                geom = "line",
                linewidth = 1,
                color = "red") +
  stat_function(fun = df, 
                args = list(df1 = 4, df2 = 45), 
                geom = "line",
                linewidth = 1,
                 color = "blue") +
stat_function(fun = df,
              args = list(df1 = 9, df2 = 40),
              geom = "line",
              linewidth = 1,
              color = "forestgreen") +
  scale_x_continuous(breaks = 0:5) +
  labs(x = "F", y = "") +
  annotate("text", 
           x = 3, y = 0.8, 
           label = "F(2,47)", 
           color = "red",
           size = 6) +
  annotate("text", 
           x = 3, y = 0.6, 
           label = "F(4,45)", 
           color = "blue",
           size = 6) +
  annotate("text",
           x = 3, y = 0.4,
           label = "F(9,40)",
           color = "forestgreen",
           size = 6)
```

* 50 subjects in 3 vs 5 vs 10 groups

## Example data

* $N$ = 57 observations on the following 2 variables.
  * `Group`
    * `mAD`=Alzheimer’s disease
    * `MCI`=mild impairment
    * `NCI`=no impairment
  * `Abeta`
    * Amount of amyloid beta from the posterior cingulate cortex (pmol/g tissue)

## Example data

```{r}
#| code-fold: true
Amyloid %>% group_by(Group) %>%
  summarize(mean = mean(Abeta),
            variance = var(Abeta)) %>%
  kable()
```


## Check assumptions: Equal variance

:::: {.columns}

::: {.column width="50%"}

* Graphically

```{r}
#| code-fold: true
ggplot(data = Amyloid,
       aes(x = Group, y = Abeta)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(width = 0.2)
```

:::

::: {.column width="50%"}

* `leveneTest()` in **car** package

```{r}
#| code-fold: true
library(car)
leveneTest(Abeta ~ Group, data = Amyloid)
```

* $H_0$: No differences in variances across groups
  * NS result = equal variances

:::

::::

## Example: ANOVA

```{r}
#| code-fold: true
m1 <- lm(Abeta ~ Group,
         contrasts=list(Group='contr.sum'),
         data = Amyloid)
anova(m1)
```

* We reject the null hypothesis that all means are equal, F(2, 54) = 5.9706, p = .0045. **At least one group mean is different from the others.**

## Type I vs III sums of squares

* **Type III sums of squares** are "corrected sums of squares"
  * How ANOVA is taught, default in every software package except R
* **Type I sums of squares** are "uncorrected sums of squares"
  * Default in R
* Different interpretations
  * But not different numerically **if all groups are the same size**
  * Type III recommended by *almost* everyone when groups are different sizes

## Assumptions: Normality

:::: {.columns}

::: {.column width="50%"}

* Graphically

```{r}
#| code-fold: true
m1resid <- data.frame(residuals(m1))
ggplot(data = m1resid, 
       aes(sample = residuals.m1.)) +
  geom_qq() +
  geom_qq_line()
```

```{r}
#| code-fold: true
m1resid <- data.frame(residuals(m1))
ggplot(data = m1resid, 
       aes(x = residuals.m1.)) +
  geom_histogram()
```

:::

::: {.column width="50%"}

* `shapiro.test()` in **stats**

```{r}
#| code-fold: true
shapiro.test(resid(m1))
```

* $H_0$: Observations come from normally distributed population
  * NS result = normally distributed residuals

:::

::::

## Post hoc tests

* Significant overall (omnibus) test for ANOVA means
  * *At least one group mean* is different from the others
  * But **which one(s)**?
* Post hoc tests
  * Figure out which means are different from one another
  * But not all possible tests, only specific ones
  
## Post hoc tests

* Conceptually, a series of $t$-tests
  * Use standard error estimate based on **whole model** instead of standard error estimate from just the two groups
  * Context of whole model
    * Not just some random set of 2 group comparisons

## Contrasts for post hoc tests

* Many built-in contrasts for specific patterns
  * Four common: dummy, deviation, polynomial, Helmert
  * Also: User-specified contrasts
* Contrasts / weights: [UCLA Statistical Consulting](https://stats.oarc.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/)
* ANOVA table is the **same** regardless of which contrasts you use
  * `summary()` of model will give you mean differences

## Example again: ANOVA

```{r}
#| code-fold: true
m1 <- lm(Abeta ~ Group,
         contrasts=list(Group='contr.sum'),
         data = Amyloid)
anova(m1)
```

## Contrasts: Dummy / `treatment`

* Compare each group to a specific group (e.g., control)

&nbsp; | mAD v MCI | mAD v NCI  
------|:------:|:------:
mAD   | 0    | 0
MCI   | 1    | 0
NCI   | 0    | 1

* Default reference: First

## Contrasts: Dummy / `treatment`

```{r}
#| code-fold: true
m1_trt <- lm(Abeta ~ Group,
         contrasts=list(Group='contr.treatment'),
         data = Amyloid)
#anova(m1_trt)
summary(m1_trt)
```

## Contrasts: Deviation / `sum`

* Compare each group (except one) to *unweighted* grand mean
  * $(\bar{X}_1 + \bar{X}_2 + \bar{X}_3)/3$

&nbsp; | mAD v GM | MCI v GM 
------|:------:|:------:
mAD   | 1    | 0     
MCI   | 0    | 1     
NCI   | -1   | -1    

* Default left out: Last 

## Contrasts: Deviation / `sum`

```{r}
#| code-fold: true
m1_sum <- lm(Abeta ~ Group,
         contrasts=list(Group='contr.sum'),
         data = Amyloid)
#anova(m1_sum)
summary(m1_sum)
```

## Contrasts: Polynomial / `poly`

* Linear, quadatric, etc. effects (orthogonal)

&nbsp; | Linear | Quadratic  
------|:------:|:------:
mAD   | -1   | 1
MCI   | 0    | -2
NCI   | 1    | 1

* Only useful with ordered levels (none, low, high)

## Contrasts: Polynomial / `poly`

```{r}
#| code-fold: true
m1_poly <- lm(Abeta ~ Group,
              contrasts=list(Group='contr.poly'),
              data = Amyloid)
#anova(m1_poly)
summary(m1_poly)
```

## Contrasts: Helmert / `helmert`

* Compare 1st to mean of 2 to k, 2nd to mean of 3 to k, etc.

&nbsp; | mAD v MCI | mAD v NCI  
------|:------:|:------:
mAD   | 2/3   | 0
MCI   | -1/3  | 1/2
NCI   | -1/3  | -1/2

## Contrasts: Helmert / `helmert`

```{r}
#| code-fold: true
m1_helm <- lm(Abeta ~ Group,
              contrasts=list(Group='contr.helmert'),
              data = Amyloid)
#anova(m1_helm)
summary(m1_helm)
```

## Mean comparisons

:::: {.columns}

::: {.column width="50%"}

```{r}
#| code-fold: true
ggplot(data = Amyloid,
       aes(x = Group, 
           y = Abeta, 
           color = Group)) +
  geom_jitter(width = 0.2) +
  geom_point(data = means, 
             aes(x = Group, 
                 y = meanAbeta, 
                 color = Group,
                 size = 10)) +
  geom_hline(yintercept = sum(means$meanAbeta)/3, 
             linetype = "dashed") +
  theme(legend.position="none")
```

:::

::: {.column width="50%"}

* Dummy: mAD v MCI, mAD v NCI
* Sum: mAD v GM, **not** MCI v GM
* Poly: Linear, **not** quadratic
* Helmert: 1 v 2&3, **not** 2 v 3

:::

::::

## Should you correct for that?

* Arguments for YES
  * You're doing multiple tests
  * Inflated type I error rate
* Arguments for NO
  * Significant overall ANOVA means "there is a difference"
  * Specific, theoretically-based comparisons
* If you do correct, use methods we've already talked about


## Welch ANOVA

* Accounts for unequal variances across groups: Heterogeneity
* `oneway.test()` in **stats** package


```{r}
oneway.test(Abeta ~ Group, data = Amyloid, var.equal = FALSE)
```

## Kruskal Wallis

* Non-parametric version of ANOVA
* `kruskal.test()` in **stats** package

```{r}
kruskal.test(Abeta ~ Group, data = Amyloid)
```


# In-class activities

## In-class activities

* Conduct an ANOVA
* Do some post hoc tests
* Compare some means

## Next week

* Last week of lecture! Woo!
* Factorial ANOVA
  * Multiple grouping variables
    * Condition **and** sex
    * Medication 1 dose **and** medication 2 dose
* Within-subjects ANOVA
  * Extension of paired $t$-test
