---
title: "Introduction to Biostatistics"
format:
  # html: 
  #   toc: true
  #   toc-location: left
  #   slide-level: 3
  #   number-sections: true
  #   toc-depth: 2
  #   self-contained: true
  #   #output-file: "Mult22-1-handout"
  # pdf: 
  #   toc: true
  #   number-sections: true
  #   #slide-level: 3
  #   toc-depth: 2
  #   shift-heading-level-by: 0
  #   number-offset: [2]
  revealjs:
    toc: false
    number-sections: true
    #slide-level: 3
    scrollable: true
    embed-resources: true
    slide-number: true
    width: 1200
knitr: 
  opts_chunk:
    echo: true
    message: false
    warning: false
---

```{r, include = FALSE, echo = FALSE, wanring = FALSE, message= FALSE}
library(tidyverse)
library(grid)
library(xtable)
library(kableExtra)
theme_set(theme_classic(base_size = 16))
set.seed(12345)
options(scipen= 100)
```

# Learning objectives

## Learning objectives

* *Describe* the logic of factorial ANOVA
  * Including main effects and interactions
* *Conduct* and *interpret* the factorial ANOVA tests
* *Conduct* post hoc tests for the factorial ANOVA

# Review

## One-way ANOVA

* a.k.a. one factor ANOVA, between-subjects ANOVA
* One *factor* (grouping variable) with *independent* groups
  * The **"between-subjects"** factor
* One continuous outcome
* Group differences on the outcome
  * Partitioned variance: Between and within (error)
  * **Ratio** of between-group variance to within-group variance

## Post hoc tests

* One-way ANOVA has 3 or more groups
  * Omnibus test: Is at least one mean different?
  * Post hoc tests to determine **which**
  * Contrasts: Dummy (treatment), sum (grand mean), polynomial

# Factorial ANOVA

## Factorial ANOVA

* Many names
  * Factorial (not specifying *how many factors*)
  * Two-way between-subjects ANOVA
  * Two-factor between-subjects ANOVA
  * $2 \times 3$ between-subjects ANOVA
  
## Partitioned variance

* Multiple factors = more partitioning
* With 2 factors (and their **interaction**)
  * Between-group variance for *factor A*
  * Between-group variance for *factor B*
  * Between-group variance for the *interaction*
  * *Residual* (within-group) variance
  
## Interactions

* Is the effect of factor A the same *at all levels of factor B*?
  * Likewise, is the effect of factor B the same *at all levels of factor A*?
* **If not, you have an interaction**
  * *Conditional* effects
  * Is there an effect of factor A? 
    * It **depends** on the level of factor B

## Example: Group, Sex, No interaction

```{r}
#| code-fold: true
Group <- c("Control", "Treatment", "Control", "Treatment")
Sex <- c("Female", "Female", "Male", "Male")
Y1 <- c(5, 10, 10, 15)
int1 <- data.frame(Group, Sex, Y1)
ggplot(data = int1, 
       aes(x = Group, 
           y = Y1, 
           group = Sex,
           color = Sex)) +
  geom_point() +
  geom_path(aes(group = Sex)) +
  ylim(0,20)
```

## Example: Group, No Sex, No interaction

```{r}
#| code-fold: true
Y2 <- c(5, 10, 5.5, 10.5)
int2 <- data.frame(Group, Sex, Y2)
ggplot(data = int2, 
       aes(x = Group, 
           y = Y2, 
           group = Sex,
           color = Sex)) +
  geom_point() +
  geom_path(aes(group = Sex)) +
  ylim(0,20)
```

## Example: No Group, Sex, No interaction

```{r}
#| code-fold: true
Y3 <- c(5, 5, 10, 10)
int3 <- data.frame(Group, Sex, Y3)
ggplot(data = int3, 
       aes(x = Group, 
           y = Y3, 
           group = Sex,
           color = Sex)) +
  geom_point() +
  geom_path(aes(group = Sex)) +
  ylim(0,20)
```

## Example: Group, Sex, interaction (!!!)

```{r}
#| code-fold: true
Y4 <- c(5, 5, 10, 20)
int4 <- data.frame(Group, Sex, Y4)
ggplot(data = int4, 
       aes(x = Group, 
           y = Y4, 
           group = Sex,
           color = Sex)) +
  geom_point() +
  geom_path(aes(group = Sex)) +
  ylim(0,20)
```
  
## Which means for which effects?

* **Main effect** of factor A
  * Means for factor A, *ignoring factor B*
* **Main effect** of factor B
  * Means for factor B, *ignoring factor A*
* Effect of interaction
  * Means of all combinations (means in plots)
* Parallels to contingency tables: **marginal values (factors)** and **joint values (interaction)**

## Assumptions

* **Two** *categorical* (grouping) variables with 2+ categories
* One *continuous* (i.e., interval or ratio) outcome variable
* Data are *randomly sampled* from the population
* Data are *independent* (within and across groups)
* Data are approximately *normally distributed* OR *sample size is large enough* for normally distributed sampling distribution
  * Residuals from model are **normally distributed**
* Population *variance* is unknown and *same in all groups*

## Hypotheses

* Factor A
  * $H_0$: All (marginal) means for factor A are equal 
* Factor B
  * $H_0$: All (marginal) means for factor B are equal 
* Interaction
  * $H_0$: No interaction is present

## ANOVA summary table

Source of variation | Sums of squares (SS) | df | Mean squares (MS) | F
------|---------|------|------|------
BetweenA | $SS_{A}$ | $k_A - 1$ | $MS_{A}$ | $F_A$
BetweenB | $SS_{B}$ | $k_B - 1$ | $MS_{B}$ | $F_B$
Interaction | $SS_{AxB}$ | $(k_A - 1)(k_B - 1)$ | $MS_{AxB}$ | $F_{AxB}$
Error | $SS_{error}$ | $N - k_A k_B$ | $MS_{error}$
Total | $SS_{total}$ | $N - 1$

* $k_A$ and $k_B$ = number of groups for each factor
* $N$ = total number of participants (observations)

## Test statistic: $F$-statistic

* $F$ statistic for **each effect**
  * $F_A = \frac{MS_{A}}{MS_{error}}$
  * $F_B = \frac{MS_{B}}{MS_{error}}$
  * $F_{AxB} = \frac{MS_{AxB}}{MS_{error}}$
* Compare to critical $F$ from a distribution with **appropriate** degrees of freedom **for the effect**
  * If the observed $F$ statistic is larger, reject $H_0$
  * Otherwise, retain $H_0$

## Example data

* `birthwt` dataset in the **MASS** package
  * `race`: mother’s race (1 = white, 2 = black, 3 = other).
  * `smoke`: smoking status during pregnancy.
  * `bwt`: birth weight in grams
  
```{r}
#| code-fold: true
library(MASS)
data(birthwt)
birthwt <- birthwt %>%
  mutate(race = as.factor(race),
         smoke = as.factor(smoke))
head(birthwt)
```

## `race` and `smoke`

:::: {.columns}

::: {.column width="50%"}

```{r}
table(birthwt$race)
table(birthwt$smoke)
```

:::

::: {.column width="50%"}

```{r}
table(birthwt$race, birthwt$smoke)

```

:::

::::

## Means and SDs of `bwt`

```{r}
#| code-fold: true
bwt_sum <- birthwt %>% 
  group_by(race, smoke) %>%
  summarize(nbwt = length(bwt),
            meanbwt = mean(bwt),
            sdbwt = sd(bwt))
kable(bwt_sum)  
```

## Means $\pm$ 1.96 standard error

```{r}
#| code-fold: true
ggplot(data = bwt_sum, 
       aes(x = race, 
           y = meanbwt, 
           group = smoke,
           color = smoke)) +
  geom_point() +
  geom_path(aes(group = smoke)) +
  geom_errorbar(aes(ymin = meanbwt - 1.96*sdbwt/sqrt(nbwt), 
                    ymax = meanbwt + 1.96*sdbwt/sqrt(nbwt)), 
                width = 0.2) +
  labs(x = "Race", 
       y = "Birth weight (g)") +
  scale_x_discrete(labels=c("1" = "White", 
                            "2" = "Black",
                            "3" = "Other")) +
  scale_color_discrete(name = "Smoker", 
                       labels = c("No", "Yes")) +
  geom_hline(yintercept = 2500, linetype = "dashed")
```

## Check assumptions: Equal variance

:::: {.columns}

::: {.column width="50%"}

* Graphically

```{r}
#| code-fold: true
ggplot(data = birthwt,
       aes(x = race, 
           y = bwt, 
           group = interaction(race, smoke))) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(width = 0.2, aes(color = smoke)) +
  geom_hline(yintercept = 2500, linetype = "dashed")
```

:::

::: {.column width="50%"}

* `leveneTest()` in **car** package

```{r}
#| code-fold: true
library(car)
leveneTest(bwt ~ race:smoke, data = birthwt)
```

* $H_0$: No differences in variances across groups
  * NS result = equal variances

:::

::::

## Example: ANOVA 1

```{r}
m1 <- lm(bwt ~ race + smoke + race*smoke,
         contrasts=list(race = contr.sum, 
                        smoke = contr.sum),
         data = birthwt)
```

* `lm()` function for **linear model**
  * `outcome ~ predictors` format
  * Effects of `race`, `smoke`, and interaction (`race*smoke`)
  * Contrasts: Sum (`contr.sum`) in `list()` function to allow one for each factor

## Example: ANOVA 2

* `Anova()` function in **car** package
  * Provide model (`m1` on last slide)
  * `type = 3` for correct type III sums of squares

```{r}
library(car)
Anova(m1, type = 3)
```

## Example: ANOVA 3

* We **reject** the null hypothesis that all `race` means are equal, F(2, 183) = 6.18, p = .0025. **At least one `race` group mean is different from the others.**

* We **reject** the null hypothesis that all `smoke` means are equal, F(1, 183) = 7.15, p = .0082. **At least one `smoke` group mean is different from the others.**

* We **retain** the null hypothesis that there is **not** an interaction, F(2, 183) = 2.25, p = .1085. **There is no interaction of `race` and `smoke`.**
  * The effect of `race` doesn't depend on `smoke` and vice versa

## Assumptions: Normality

:::: {.columns}

::: {.column width="50%"}

* Graphically

```{r}
#| code-fold: true
m1resid <- data.frame(residuals(m1))
ggplot(data = m1resid, 
       aes(sample = residuals.m1.)) +
  geom_qq() +
  geom_qq_line()
```

```{r}
#| code-fold: true
m1resid <- data.frame(residuals(m1))
ggplot(data = m1resid, 
       aes(x = residuals.m1.)) +
  geom_histogram()
```

:::

::: {.column width="50%"}

* `shapiro.test()` in **stats**

```{r}
#| code-fold: true
shapiro.test(resid(m1))
```

* $H_0$: Observations come from normally distributed population
  * NS result = normally distributed residuals

:::

::::

## `race`, `smoke`, no interaction

```{r}
#| code-fold: true
ggplot(data = bwt_sum, 
       aes(x = race, 
           y = meanbwt, 
           group = smoke,
           color = smoke)) +
  geom_point() +
  geom_path(aes(group = smoke)) +
  geom_errorbar(aes(ymin = meanbwt - 1.96*sdbwt/sqrt(nbwt), 
                    ymax = meanbwt + 1.96*sdbwt/sqrt(nbwt)), 
                width = 0.2) +
  labs(x = "Race", 
       y = "Birth weight (g)") +
  scale_x_discrete(labels=c("1" = "White", 
                            "2" = "Black",
                            "3" = "Other")) +
  scale_color_discrete(name = "Smoker", 
                       labels = c("No", "Yes"))
```

## Raw means

:::: {.columns}

::: {.column width="50%"}

* `smoke` means

```{r}
#| code-fold: true
birthwt %>% group_by(smoke) %>%
  summarize(meanbwt = mean(bwt))

```

:::

::: {.column width="50%"}

* `race` means

```{r}
#| code-fold: true
birthwt %>% group_by(race) %>%
  summarize(meanbwt = mean(bwt))
```

:::

::::


## Main effects vs simple effects

* Main effects
  * **Main effect** of factor A, *ignoring factor B*
    * e.g., Effect of `race`, ignoring `smoke`
  * **Any time there's a significant main effect**
* Simple (main) effects
  * **Simple effect** of factor A (*at a level of factor B*)
    * e.g., Effect of `race` **for smokers** (i.e., `smoke` = 1)
  * **Only when you have a significant interaction**

## Post hoc tests: emmeans

* `emmeans()` function in **emmeans** package
  * Provide model name (`m1`)
  * Type of contrast (`spec = trt.vs.ctrl`)
    * Other options: `pairwise`, `trt.vs.ctrlk` (last group = ctl)
  * Factor for contrasts: here, only `race`
  * Adjustment for multiple comparisons: here, `"none"`
    * Other options: `tukey`, `bonf`, `fdr`

## Post hoc tests: emmeans

* Note that these are not the **raw means**
  * **Estimated marginal means** (emmeans): Model based

```{r}
library(emmeans)
emmeans(m1, spec = trt.vs.ctrl ~ race, adjust = "none")
```

## Caution re: causality

* Often, when you design an experiment that is analyzed via ANOVA, you've **randomized** people to groups
  * That allows for robust **causal inference**
    * Group differences are **causing** differences in the outcome
* People **aren't** randomized to `race` or `smoke`
  * You **cannot** say that `race` is **causing** lower birth weight
  * It is *related to* or *associated with* lower birth weight

## Limitations

* Only allows for **categorical** (grouping) predictors
  * If you have **continuous** predictors too, move to **regression models**
    * Linear (`lm()`), logistic, Poisson, gamma (`glm()`)
  * Example: Mother age, mother weight
* Using a regression approach also avoids dealing with ANOVA issues
  * Instead of contrasts, direct control over coding of predictors
    * Dummy codes

# Within-subjects ANOVA

## Within-subjects ANOVA

* Also called: repeated-measures ANOVA
* Extension of matched-pairs $t$-test to 3+ **related** groups
  * Typically repeated measures of the same unit
* No **between-groups** effects: **No groups**
  * Between-**subject** variance
  * Within-**subject** variance
  * Error variance

## Limitations of WS ANOVA

* Doesn't allow for **different #s of repeated measures**
  * Drop entire subject if even one is missing
* Often unrealistic assumptions
  * **Sphericity** / compound symmetry 
  * Incorrect type I error, reduced power if not corrected
* Like other ANOVAs, don't allow for continuous predictors
  * Continuous **time**

## Advantages of RM designs

* In general, more **powerful** than between-subjects designs
  * Each person acts as their own control
* Can examine **change** in an individual over time
  * Longitudinal data

## More extensions: Mixed ANOVA

* 1+ between-subjects factor **and** 1+ within-subjects factor
  * Group (BS) **and** time (WS)
* More complex, many of the same limitations of WS ANOVA
  * Assumption of sphericity
  * No continuous predictors
  * No different #s of repeated measures

## More extensions: Mixed models

* Regression extension for repeated measures
  * Within-subject and between-subject predictors
  * Continuous and categorical predictors
  * Different # of repeated observations per unit
  * No unrealistic assumptions (i.e., sphericity)
* More complex
  * Need to understand regression, plus more unique complexities

# In-class activities

## In-class activities

* Do a factorial ANOVA
* Look at how you can do the ANOVA wrong 
  * Type I SS (wrong) vs type III SS (correct)
* Play around a bit more with post hoc tests
  * Pairwise, treatment with last group as reference

## Next week

* Discussion of special topics
  * 3 things that you found interesting and/or helpful about the topic
    * From the articles in the list or others you've found
    * Remember that you should use the listed readings and **at least 3 more** articles

