---
title: "Introduction to Biostatistics"
format:
  # html: 
  #   toc: true
  #   toc-location: left
  #   slide-level: 3
  #   number-sections: true
  #   toc-depth: 2
  #   self-contained: true
  #   #output-file: "Mult22-1-handout"
  # pdf: 
  #   toc: true
  #   number-sections: true
  #   #slide-level: 3
  #   toc-depth: 2
  #   shift-heading-level-by: 0
  #   number-offset: [2]
  revealjs:
    toc: false
    number-sections: true
    #slide-level: 3
    scrollable: true
    self-contained: true
    slide-number: true
    width: 1200
knitr: 
  opts_chunk:
    echo: true
    message: false
    warning: false
---

```{r, include = FALSE, echo = FALSE, wanring = FALSE, message= FALSE}
library(tidyverse)
library(gapminder)
library(DescTools)
theme_set(theme_classic(base_size = 16))
set.seed(12345)
```

# Learning objectives

## Learning objectives

* Understand the different **types** of variables that can be measured
* Think about variables in terms of **central tendency** or **location**
* Think about variables in terms of **dispersion** or **spread**
* Start to think about variable **distributions** and **plots**

# Types of variables

## Types of variables

* There are a lot of ways to talk about variables
  * Continuous vs categorical
  * Quantitative vs qualitative
  * Numeric vs non-numeric
* Often loose, ambiguous, and arbitrary

## Levels of measurement

* Four **ordered** levels of measurement based on the **mathematical** operations that can be performed
  * Nominal
  * Ordinal
  * Interval
  * Ratio

> Stevens, S. S. (1946). On the theory of scales of measurement. Science, 103(2684), 677-680.
  
## Nominal variables

* **Categories** with **no intrinsic ordering**
  * Nominal = "name"
* Examples
  * Department: Psychology, Epidemiology, Statistics, Business
  * Religion: Christian, Jewish, Muslim, Atheist
  * Ice cream flavor: vanilla, chocolate, strawberry

## Ordinal variables

* **Categories** with **some intrinsic ordering**
  * Ordinal = "ordered"
  * Differences between categories are **not meaningful/equal**
* Examples
  * Dose of treatment: low, medium, high
  * Rankings: 1st, 2nd, 3rd, 4th
  * Education: high school, some college, college grad, graduate
  * Likert scales: agree, neutral, disagree

## Interval variables

* **Quantitative variables** with **no meaningful 0 point**
  * (“Meaningful 0”: value of 0 = nothing)
  * **Differences** between values are meaningful but **ratios** are not!
* Example: Temperature in Fahrenheit or Celsius
  * **Difference** from 100F to 90F = **difference** from 90F to 80F
  * But 100F is **not twice** 50F (because 0F is arbitrary)
* Most “continuous” variables you deal with are **interval**
  * Most statistical procedures assume interval-level measurement
  
## Ratio variables

* **Quantitative variables** with **meaningful 0 point**
  * (“Meaningful 0”: value of 0 = nothing)
  * **Differences** between values are meaningful and so are **ratios**!
* Example: Temperature in Kelvin
  * **Difference** from 100K to 90K = **difference** from 90K to 80K
  * 100K is **twice as hot** as 50K (0K is *zero* molecular movement)
* Height, weight, age, counts

## Stevens (1946)

The *levels of measurement* determine what **mathematical** (and **statistical**) operations you can perform

Mathematical operation | Nominal | Ordinal | Interval | Ratio
-------------------------|:------:|:------:|:------:|:------:
equal, not equal | $\checkmark$ | $\checkmark$ | $\checkmark$ | $\checkmark$
greater or less than |  | $\checkmark$ | $\checkmark$ | $\checkmark$
add, subtract |  |  | $\checkmark$ | $\checkmark$
multiply, divide |  |  |  | $\checkmark$
central tendency | mode | median | mean | mean

## **R** "translations" of variable types

* Nominal
  * `char` = "character": Text variable, also called "string"
* Nominal or ordinal
  * `fctr` = "factor": Ordered or unordered categories
  * `log` = "logical" or "boolean": `TRUE` or `FALSE`
* Interval or ratio
  * `int` = "integer": Whole number (no decimals)
  * `dbl` = "double" (`num`): Number w possible decimal places

# Central tendency

## Central tendency

* Also called "location"
  * Where are the observations *located*?
* Several measures of central tendency
  * Depend on scale type
  * Mode, median, mean

## Mode

* The most frequently occurring value
  * Value actually exists in the data
* Used for *nominal*, *ordinal*
  * Can also be used for *interval*, and *ratio* variables but less useful

## Median

* Order observations from smallest to largest 
  * Select the **middle** observation
  * Value actually exists in the data (if $n$ is even)
* Used for *ordinal*, *interval*, and *ratio* variables
  
## Mean

* Add up all observations and divide by the number of observations
  * *Value may not actually occur in the data*
* Mathematically: $\frac{\sum_{i=1}^n X_i}{n}$
* Used for *interval* and *ratio* variables

## **gapminder** dataset

```{r}
gap <- gapminder
head(gap)
```

## **gapminder** variables

Variable | R variable type | Scale type
---------|-----------------|-----------
`country`       | factor            | nominal
`continent`     | factor            | nominal
`year`          | integer           | ordinal / interval
`lifeExp`       | double / numeric  | ratio
`pop`           | integer           | ratio (count)
`gdpPercap`     | double / numeric  | ratio

## Nominal variables: Mode

* `Mode()` function from **DescTools** package

```{r}
Mode(gap$country)
```

* There is *no single mode* for `country`

## Nominal variables: Mode

* `Mode()` function from **DescTools** package

```{r}
Mode(gap$continent)
```

* There are 52 countries (52 countries * 12 years = 624) in Africa
  * The value of "Africa" is the **mode** of the `continent` variable

## Ordinal variables: Median

```{r}
median(gap$year, na.rm = TRUE)
```

* <span style="color:red">1952, 1957, 1962, 1967, 1972, 1977,</span> <span style="color:blue">1982, 1987, 1992, 1997, 2002, 2007</span>
  * Median is halfway between 1977 and 1982
* `na.rm = TRUE` option removes missing values
  * If you have missing values and don't include this, you will get `NA`

## Interval variables: Mean

```{r}
mean(gap$year, na.rm = TRUE)
```

* $(1952 + 1957 + \dots + 2002 + 2007)/12$
* You can also calculate the median of an interval variable, but we just did that for this variable (treating it as ordinal)
* `na.rm = TRUE` option removes missing values
  * If you have missing values and don't include this, you will get `NA`

## Use only data from 2007

* Including all data gives us a confusing mish-mash of information
  * I'm just going to include data from 2007 for these variables
* Also going to create the `gdp` variable at the same time

```{r}
gap_2007 <- gapminder %>% 
  mutate(gdp = gdpPercap * pop) %>%
  filter(year == 2007)
```

## Ratio variables: Median

```{r}
median(gap_2007$lifeExp, na.rm = TRUE)
median(gap_2007$pop, na.rm = TRUE)
median(gap_2007$gdpPercap, na.rm = TRUE)
median(gap_2007$gdp, na.rm = TRUE)
```

* 50% of countries have a life expectancy **less than** 71.9355 years 
* 50% of countries have a life expectancy **greater than** 71.9355 years

## Ratio variables: Mean

```{r}
mean(gap_2007$lifeExp, na.rm = TRUE)
mean(gap_2007$pop, na.rm = TRUE)
mean(gap_2007$gdpPercap, na.rm = TRUE)
mean(gap_2007$gdp, na.rm = TRUE)
```

* The average life expectancy is 67.00742 years

## Compare mean and median

* Mean and median are both measures of *central tendency* for numbers
  * **What does it mean when they differ?**
* Median involves only values near it
  * Extreme values don't impact it
* Mean involves **all** values
  * **Extreme values impact the mean**
  * Mean is "pulled toward" extreme values

## Mean and median of `lifeExp`

```{r}
#| code-fold: true
ggplot(data = gap_2007, aes(x = lifeExp)) + 
  geom_histogram(bins = 30, color = "grey80") +
  geom_vline(xintercept = mean(gap_2007$lifeExp), color = "blue", linewidth = 1.5) +
  geom_vline(xintercept = median(gap_2007$lifeExp), color = "red", linewidth = 1.5, linetype = "dashed")
```

## Mean and median of `pop`

```{r}
#| code-fold: true
ggplot(data = gap_2007, aes(x = pop)) + 
  geom_histogram(bins = 30, color = "grey80") +
  geom_vline(xintercept = mean(gap_2007$pop), color = "blue", linewidth = 1.5) +
  geom_vline(xintercept = median(gap_2007$pop), color = "red", linewidth = 1.5, linetype = "dashed")
```

## Mean and median of `gdpPercap`

```{r}
#| code-fold: true
ggplot(data = gap_2007, aes(x = gdpPercap)) + 
  geom_histogram(bins = 30, color = "grey80") +
  geom_vline(xintercept = mean(gap_2007$gdpPercap), color = "blue", linewidth = 1.5) +
  geom_vline(xintercept = median(gap_2007$gdpPercap), color = "red", linewidth = 1.5, linetype = "dashed")
```

<!-- ## Mean and median of `gdp` in 2007 -->

<!-- ```{r} -->
<!-- ggplot(data = gap_2007, aes(x = gdp)) +  -->
<!--   geom_histogram(bins = 30, color = "grey80") + -->
<!--   geom_vline(xintercept = mean(gap_2007$gdp), color = "blue", linewidth = 1.5) + -->
<!--   geom_vline(xintercept = median(gap_2007$gdp), color = "red", linewidth = 1.5, linetype = "dashed") -->
<!-- ``` -->

# Dispersion

## Dispersion

* Also called "spread"
  * How spread out or *dispersed* are the observations?
  * Only applies to numeric (interval or ratio) variables
* Several measures of dispersion
  * Depend on scale type and variable distribution
  * Minimum and maximum, percentiles, standard deviation

## Minimum and maximum

:::: {.columns}

::: {.column width="50%"}

* Minimum

```{r}
min(gap_2007$lifeExp, na.rm = TRUE)
min(gap_2007$pop, na.rm = TRUE)
min(gap_2007$gdpPercap, na.rm = TRUE)
```

:::

::: {.column width="50%"}

* Maximum

```{r}
max(gap_2007$lifeExp, na.rm = TRUE)
max(gap_2007$pop, na.rm = TRUE)
max(gap_2007$gdpPercap, na.rm = TRUE)
```

:::

::::

## Percentiles 

* Value **below which** some percentage of observations lie
  * 67th percentile: 67% of observations have values below this value
  * Median is the 50th percentile: 50% below (and 50% above)
  * Quartiles divide into 4 parts: 25th, 50th, 75th percentiles
  * Standardized tests: 97th percentile means you scored higher than 97% of people taking the test
  
## Percentiles

```{r}
quantile(gap_2007$lifeExp, c(0.25, 0.5, 0.75))
quantile(gap_2007$pop, c(0.25, 0.5, 0.75))
quantile(gap_2007$gdpPercap, c(0.1, 0.9))
```

## Standard deviation (& variance)

* Spread around the mean
* Same units as the variable 
  * Unlike variance, with is $SD^2$
* Influenced by extreme values
* Mathematically: $\sqrt{\frac{\sum (X_i - \overline{X})^2}{n-1}}$

## Presenting location & dispersion

* Symmetric distribution
  * Mean $\pm$ standard deviation
  * Median (25th to 75th percentile)
  * Median (min, max)
* Asymmetric distribution
  * Median (25th to 75th percentile)
  * Median (min, max)

## Presenting

```{r}
#| code-fold: true
lifeexp_perc <- quantile(gap_2007$lifeExp, c(0, 0.25, 0.5, 0.75, 1))
lifeexp_sd <- sd(gap_2007$lifeExp, na.rm = TRUE)
lifeexp_mean <- mean(gap_2007$lifeExp, na.rm= TRUE)
ggplot(data = gap_2007, aes(x = lifeExp)) + 
  geom_histogram(bins = 30, color = "grey80")
```

## Presenting: Mean plus/minus 1 SD

```{r}
#| code-fold: true
ggplot(data = gap_2007, aes(x = lifeExp)) + 
  geom_histogram(bins = 30, color = "grey80") +
  annotate("pointrange", x = lifeexp_mean, y = 5, 
            xmin = mean(gap_2007$lifeExp) - lifeexp_sd,
            xmax = mean(gap_2007$lifeExp) + lifeexp_sd,
            size = 1.5, linewidth = 1.5, color = "red") +
  ylab("count")
```

## Presenting: Median, 25th, 75th %iles

```{r}
#| code-fold: true
ggplot(data = gap_2007, aes(x = lifeExp)) + 
  geom_histogram(bins = 30, color = "grey80") +
  annotate("pointrange", x = lifeexp_perc[3], y = 5, 
            xmin = lifeexp_perc[2],
            xmax = lifeexp_perc[4],
            size = 1.5, linewidth = 1.5, color = "red") +
  ylab("count")
```

# In-class activities

## In-class activities

* Look at some data and **variables**
  * Think about what **type(s)** of variables they are
* Examine **central tendency** and **dispersion** for the variables
  * Present the **best values** to describe the data
* Start to think about variable **distributions** and **plots**

