---
title: "BTS 510 Lab 4"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
set.seed(12345)
library(tidyverse)
library(Stat2Data)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* Select an **appropriate plot** for the **variable type**
* Create plots in **ggplot2** 

## Data 

* `ICU` data from the **Stat2Data** package
  * `ID`: Patient ID code
  * `Survive`: 1 = patient survived to discharge or 0 = patient died
  * `Age`: Age (in years)
  * `AgeGroup`: 1 = young (under 50), 2 = middle (50-69), 3 = old (70+)
  * `Sex`: 1 = female or 0 = male
  * `Infection`: 1 = infection suspected or 0 = no infection
  * `SysBP`: Systolic blood pressure (in mm of Hg)
  * `Pulse`: Heart rate (beats per minute)
  * `Emergency`: 1 = emergency admission or 0 = elective admission

* Convert the factor variables to factor variables, as in the lecture
  * `as.factor()` function

```{r}
library(Stat2Data)
data(ICU)
nrow(ICU)
ICU$Survive <- as.factor(ICU$Survive)
ICU$Sex <- as.factor(ICU$Sex)
ICU$Infection <- as.factor(ICU$Infection)
ICU$Emergency <- as.factor(ICU$Emergency)
ICU$AgeGroup <- as.factor(ICU$AgeGroup)
```


## Tasks

1. Make a histogram of blood pressure. Make the bars grey with a black outline. Add vertical lines at the standard cutoffs ([https://newsroom.heart.org/news/high-blood-pressure-redefined-for-first-time-in-14-years-130-is-the-new-high](https://newsroom.heart.org/news/high-blood-pressure-redefined-for-first-time-in-14-years-130-is-the-new-high)) of 120, 130, and 140. Make those lines green, yellow, and red, respectively.

```{r}
plot1a <- ggplot(data = ICU, aes(x = SysBP)) +
    geom_histogram(color = "black", fill = "grey")
plot1a

plot1b <- plot1a +
    geom_vline(xintercept = 120, color = "green", linewidth = 1.5) +
    geom_vline(xintercept = 130, color = "yellow", linewidth = 1.5) +
    geom_vline(xintercept = 140, color = "red", linewidth = 1.5)
plot1b
```


2. Make dotplots of blood pressure for emergency vs elective admission patients. Try different numbers of bins or binwidths. 

```{r}
plot2a <- ggplot(data = ICU, aes(x = Emergency, y = SysBP)) +
    geom_dotplot(binwidth = 1,
                 method = "histodot",
                 binaxis = "y",
                 stackdir = "center")
plot2b <- ggplot(data = ICU, aes(x = Emergency, y = SysBP)) +
    geom_dotplot(binwidth = 5,
                 method = "histodot",
                 binaxis = "y",
                 stackdir = "center")
plot2c <- ggplot(data = ICU, aes(x = Emergency, y = SysBP)) +
    geom_dotplot(binwidth = 10,
                 method = "histodot",
                 binaxis = "y",
                 stackdir = "center")
plot2a
plot2b
plot2c
```


3. Make a scatterplot of blood pressure (Y) vs age (X). Add a straight line to the plot. Does it look like blood pressure increases, decreases, or is relatively stable over ages?

```{r}
ggplot(data = ICU, aes(x = Age, y = SysBP)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE)
ggplot(data = ICU, aes(x = Age, y = SysBP)) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE)
```

