---
title: "BTS 510 Lab 6"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
library(tidyverse)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* Understand **probability** as a measure of **uncertainty**
* **Summarize** probabilities with *probability distributions* 
* Describe the difference between **population** and **sample**
* Think about **uncertainty** in sampling
* **Summarize** uncertainty in sampling: **Sampling distribution**

## Mathematically-derived sampling distribution

* Many combinations of **population distribution** and **test statistic** have *mathematically-derived* sampling distributions
  * For example, with a **normally distributed** population ($\mathcal{N}(\mu, \sigma^2)$), the sampling distribution of the **mean** is $\bar{X} \sim \mathcal{N}(\mu, \frac{\sigma^2}{n})$
    * In this case, the *population* distribution and *sampling* distribution are the **same distribution**: Normal
    * The variance (or standard deviation) of the sampling distribution **decreases** as $n$ increases: A *larger sample* leads to a smaller, *more precise standard error*

::: {.callout-note}

* There is not a mathematically-derived sampling distribution for every population distribution and test statistic
  * If you're dealing with *unknown or unusual distributions or test statistics*, your only option may be *computational methods*
  
:::

## Computationally-derived sampling distribution

### Getting started

* Set a **random number seed**
  * If you don't do this, every time you run the code, you'll get different numbers
  * Setting a random seed makes your simulation *replicable*
  * (I normally set a seed at the start of each document, but I moved it down here because it's very important in this lab.)
* Load (and install, if needed) the [**infer** package](https://infer.tidymodels.org/)
  * **infer** has a function that we'll use to *repeatedly sample* 
  * There are other ways to do this, but this is very simple and works well
* Only have to do this once per document

```{r}
set.seed(12345)
library(infer)
```


### Standard Normal population: $population \sim \mathcal{N}(0, 1)$

#### Create a population

* `rnorm()` creates a random variable that follows a **normal** distribution 
  * First argument = Number of observations (population size)
  * Second argument = $\mu$
  * Third argument = $\sigma$ (not $\sigma^2$)

```{r}
norm_pop <- rnorm(1000000, 0, 1)
mean(norm_pop)
var(norm_pop)
```

#### Sample from the population

* Small sample: $n$ = 10
  * `rep_sample_n()` repeatedly samples from the dataset
    * Piped (`%>%`) from the `norm_pop` dataset, so it resamples from that
    * `size = 10`: Sample size for each sample
    * `reps = 10000`: Number of repeated samples (could be larger but this is good enough to show how things work and doesn't take too long to run)
    * `replace = TRUE`: Sample with replacement
  * `summarise(x_bar = mean(norm_pop))`: Create a data frame with each of the means from each sample

```{r}
norm_means10 <- data.frame(norm_pop) %>%
  rep_sample_n(size = 10, reps = 10000, replace = TRUE) %>%
  summarise(x_bar = mean(norm_pop))

ggplot(data = norm_means10, aes(x = x_bar)) +
  geom_histogram(bins = 50) +
  xlim(-1.5, 1.5)
```

* Moderate sample: $n$ = 50

```{r}
norm_means50 <- data.frame(norm_pop) %>%
  rep_sample_n(size = 50, reps = 10000, replace = TRUE) %>%
  summarise(x_bar = mean(norm_pop))

ggplot(data = norm_means50, aes(x = x_bar)) +
  geom_histogram(bins = 50) +
  xlim(-1.5, 1.5) 
```

* Large sample: $n$ = 100

```{r}
norm_means100 <- data.frame(norm_pop) %>%
  rep_sample_n(size = 100, reps = 10000, replace = TRUE) %>%
  summarise(x_bar = mean(norm_pop))

ggplot(data = norm_means100, aes(x = x_bar)) +
  geom_histogram(bins = 50) +
  xlim(-1.5, 1.5)
```

::: {.callout-note}

* Note that I adjusted the X axis to be the same for all three plots using `xlim()` so that you can easily compare the distributions 
* You'll want to do the same in your later plots too
  
:::

#### Compare estimates

* Compare means and variances for each sample size

```{r}
mean(norm_means10$x_bar)
mean(norm_means50$x_bar)
mean(norm_means100$x_bar)
```

```{r}
var(norm_means10$x_bar)
var(norm_means50$x_bar)
var(norm_means100$x_bar)
```

* How does each value correspond to what you'd expect based on the mathematically-derived sampling distribution?
  * $\bar{X} \sim \mathcal{N}(\mu, \frac{\sigma^2}{n})$

### Bernoulli population: $population \sim B(0.3)$

#### Create a population

* `rbinom()` creates a random variable that follows a **binomial** distribution 
  * First argument = Number of observations (population size)
  * Second argument = number of trials (here, just 1 for a **Bernoulli** trial)
  * Third argument = $p$ (probability of success)

```{r}
bernoulli_pop <- rbinom(1000000, 1, 0.3)
mean(bernoulli_pop)
var(bernoulli_pop)
```

#### Sample from the population

* Small sample: $n$ = 10

```{r}
bern_means10 <- data.frame(bernoulli_pop) %>%
  rep_sample_n(size = 10, reps = 10000, replace = TRUE) %>%
  summarise(x_bar = mean(bernoulli_pop))

ggplot(data = bern_means10, aes(x = x_bar)) +
  geom_histogram(bins = 50) +
    xlim(0,1)
```

* Moderate sample: $n$ = 50

```{r}
bern_means50 <- data.frame(bernoulli_pop) %>%
  rep_sample_n(size = 50, reps = 10000, replace = TRUE) %>%
  summarise(x_bar = mean(bernoulli_pop))

ggplot(data = bern_means50, aes(x = x_bar)) +
  geom_histogram(bins = 50) +
    xlim(0,1)
```

* Large sample: $n$ = 100

```{r}
bern_means100 <- data.frame(bernoulli_pop) %>%
  rep_sample_n(size = 100, reps = 10000, replace = TRUE) %>%
  summarise(x_bar = mean(bernoulli_pop))

ggplot(data = bern_means100, aes(x = x_bar)) +
  geom_histogram(bins = 50) +
    xlim(0,1)
```

#### Compare estimates

* Compare means and variances for each sample size

```{r}
mean(bern_means10$x_bar)
mean(bern_means50$x_bar)
mean(bern_means100$x_bar)
```

```{r}
var(bern_means10$x_bar)
var(bern_means50$x_bar)
var(bern_means100$x_bar)
```

### Binomial population: $population \sim Bin(50, 0.3)$

#### Create a population

* `rbinom()` creates a random variable that follows a **binomial** distribution 
  * First argument = Number of observations (population size)
  * Second argument = $m$ (number of trials)
  * Third argument = $p$ (probability of success)

```{r}
binomial_pop <- rbinom(1000000, 50, 0.3)
mean(binomial_pop)
var(binomial_pop)
```

#### Sample from the population

* Small sample: $n$ = 10

```{r}
bin_means10 <- data.frame(binomial_pop) %>%
  rep_sample_n(size = 10, reps = 10000, replace = TRUE) %>%
  summarise(x_bar = mean(binomial_pop))

ggplot(data = bin_means10, aes(x = x_bar)) +
  geom_histogram(bins = 50) +
    xlim(10,20)
```

* Moderate sample: $n$ = 50

```{r}
bin_means50 <- data.frame(binomial_pop) %>%
  rep_sample_n(size = 50, reps = 10000, replace = TRUE) %>%
  summarise(x_bar = mean(binomial_pop))

ggplot(data = bin_means50, aes(x = x_bar)) +
  geom_histogram(bins = 50) +
    xlim(10,20)
```

* Large sample: $n$ = 100

```{r}
bin_means100 <- data.frame(binomial_pop) %>%
  rep_sample_n(size = 100, reps = 10000, replace = TRUE) %>%
  summarise(x_bar = mean(binomial_pop))

ggplot(data = bin_means100, aes(x = x_bar)) +
  geom_histogram(bins = 50) +
    xlim(10,20)
```

#### Compare estimates

* Compare means and variances for each sample size

```{r}
mean(bin_means10$x_bar)
mean(bin_means50$x_bar)
mean(bin_means100$x_bar)
```

```{r}
var(bin_means10$x_bar)
var(bin_means50$x_bar)
var(bin_means100$x_bar)
```

### Poisson population: $population \sim Poisson(1.5)$

#### Create a population

* `rpois()` creates a random variable that follows a **Poisson** distribution 
  * First argument = Number of observations (population size)
  * Second argument = $\lambda$ (mean and variance)

```{r}
poisson_pop <- rpois(1000000, 1.5)
mean(poisson_pop)
var(poisson_pop)
```

#### Sample from the population

* Small sample: $n$ = 10

```{r}
poi_means10 <- data.frame(poisson_pop) %>%
  rep_sample_n(size = 10, reps = 10000, replace = TRUE) %>%
  summarise(x_bar = mean(poisson_pop))

ggplot(data = poi_means10, aes(x = x_bar)) +
  geom_histogram(bins = 50) +
    xlim(0,5) 
```

* Moderate sample: $n$ = 50

```{r}
poi_means50 <- data.frame(poisson_pop) %>%
  rep_sample_n(size = 50, reps = 10000, replace = TRUE) %>%
  summarise(x_bar = mean(poisson_pop))

ggplot(data = poi_means50, aes(x = x_bar)) +
  geom_histogram(bins = 50) +
    xlim(0,5)  
```

* Large sample: $n$ = 100

```{r}
poi_means100 <- data.frame(poisson_pop) %>%
  rep_sample_n(size = 100, reps = 10000, replace = TRUE) %>%
  summarise(x_bar = mean(poisson_pop))

ggplot(data = poi_means100, aes(x = x_bar)) +
  geom_histogram(bins = 50) +
    xlim(0,5)  
```

#### Compare estimates

* Compare means and variances for each sample size

```{r}
mean(poi_means10$x_bar)
mean(poi_means50$x_bar)
mean(poi_means100$x_bar)
```

```{r}
var(poi_means10$x_bar)
var(poi_means50$x_bar)
var(poi_means100$x_bar)
```