---
title: "BTS 510 Lab 9"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
library(tidyverse)
set.seed(12345)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* *Interpret* tests comparing **two unrelated samples**
* *Summarize data* using **contingency tables**
* *Describe* different study designs for **contingency tables** 

## Data 

* `Pulse` dataset from the **Stat2Data** package
    * A dataset with $n$ = 232 observations on the following 7 variables.
        * `Active`: Pulse rate (beats per minute) after exercise
        * `Rest`: Resting pulse rate (beats per minute)
        * `Smoke`: 1=smoker or 0=nonsmoker
        * `Sex`: 1=female or 0=male
        * `Exercise`: Typical hours of exercise (per week)
        * `Hgt`: Height (in inches)
        * `Wgt`: Weight (in pounds)

## Tasks

* Make plots of variables as needed (e.g., to assess assumptions)
* Conduct a $z$-test, $t$-test, and Welch's $t$-test
  * What is/are your conclusion(s) based on the tests?
  * Are the assumptions met?
    * e.g., large enough sample to justify $z$ test using sample variance
    * e.g., equal variances in both groups
  * Which test seems the best choice? (Don't make this decision based on what is significant -- here or elsewhere)
    * Do you think a non-parametric test might be a good option?

### Some useful code

* To split the dataset into `Smoke` = 0 and `Smoke` = 1
  * There are other ways to do this, so you don't *need* to use this code

```{r}
library(Stat2Data)
data(Pulse)
library(tidyverse)
Pulse_smoke <- Pulse %>% filter(Smoke == 1)
Pulse_nosmoke <- Pulse %>% filter(Smoke == 0)
head(Pulse_smoke)
head(Pulse_nosmoke)
```

* Use `alternative = "greater"` if $H_1$: $\mu_1 > \mu_2$
  * Use `alternative = "less"` if $H_1$: $\mu_1 < \mu_2$
  * Where $\mu_1$ is the mean for the first-entered group (`x`)
  * The order you enter them (`x` vs `y`) doesn't matter, **just make sure you set up the directional hypothesis accordingly**

### Active pulse rate

```{r}
mean(Pulse_smoke$Active)
mean(Pulse_nosmoke$Active)
var(Pulse_smoke$Active)
var(Pulse_nosmoke$Active)
ggplot(data = Pulse_nosmoke, 
       aes(x = Active)) +
  geom_histogram(fill = "red", 
                 alpha = 0.5, 
                 bins = 30) +
  geom_histogram(data = Pulse_smoke, 
                 aes(x = Active), 
                 fill = "black", 
                 bins = 30)
```

* Is active pulse rate **higher** among smokers than non-smokers?

```{r}
library(BSDA)
ztest1 <- z.test(x = Pulse_smoke$Active, 
                y = Pulse_nosmoke$Active, 
                sigma.x = sd(Pulse_smoke$Active), 
                sigma.y = sd(Pulse_nosmoke$Active), 
                alternative = "greater")
ztest1
ttest1 <- t.test(x = Pulse_smoke$Active, 
                y = Pulse_nosmoke$Active, 
                alternative = "greater",
                var.equal = TRUE)
ttest1
ttest1b <- t.test(x = Pulse_smoke$Active, 
                y = Pulse_nosmoke$Active, 
                alternative = "greater",
                var.equal = FALSE)
ttest1b
```

### Weight

```{r}
mean(Pulse_smoke$Wgt)
mean(Pulse_nosmoke$Wgt)
var(Pulse_smoke$Wgt)
var(Pulse_nosmoke$Wgt)
ggplot(data = Pulse_nosmoke, 
       aes(x = Wgt)) +
  geom_histogram(fill = "red", 
                 alpha = 0.5, 
                 bins = 30) +
  geom_histogram(data = Pulse_smoke, 
                 aes(x = Wgt), 
                 fill = "black", 
                 bins = 30)
```

* Do smokers weight **less than** non-smokers?

```{r}
ztest2 <- z.test(x = Pulse_smoke$Wgt, 
                y = Pulse_nosmoke$Wgt, 
                sigma.x = sd(Pulse_smoke$Wgt), 
                sigma.y = sd(Pulse_nosmoke$Wgt), 
                alternative = "less")
ztest2
ttest2 <- t.test(x = Pulse_smoke$Wgt, 
                y = Pulse_nosmoke$Wgt, 
                alternative = "less",
                var.equal = TRUE)
ttest2
ttest2b <- t.test(x = Pulse_smoke$Wgt, 
                y = Pulse_nosmoke$Wgt, 
                alternative = "less",
                var.equal = FALSE)
ttest2b
```

### Exercise

```{r}
mean(Pulse_smoke$Exercise)
mean(Pulse_nosmoke$Exercise)
var(Pulse_smoke$Exercise)
var(Pulse_nosmoke$Exercise)
ggplot(data = Pulse_nosmoke, 
       aes(x = Exercise)) +
  geom_histogram(fill = "red", 
                 alpha = 0.5, 
                 bins = 30) +
  geom_histogram(data = Pulse_smoke, 
                 aes(x = Exercise), 
                 fill = "black", 
                 bins = 30)
```

* Do smokers and non-smokers exercise the **same** amount?

```{r}
ztest3 <- z.test(x = Pulse_smoke$Exercise, 
                y = Pulse_nosmoke$Exercise, 
                sigma.x = sd(Pulse_smoke$Exercise), 
                sigma.y = sd(Pulse_nosmoke$Exercise), 
                alternative = "two.sided")
ztest3
ttest3 <- t.test(x = Pulse_smoke$Exercise, 
                y = Pulse_nosmoke$Exercise, 
                alternative = "two.sided",
                var.equal = TRUE)
ttest3
ttest3b <- t.test(x = Pulse_smoke$Exercise, 
                y = Pulse_nosmoke$Exercise, 
                alternative = "two.sided",
                var.equal = FALSE)
ttest3b
```

```{r}
library(coin)
median_test(Exercise ~ as.factor(Smoke), data = Pulse)
wilcox.test(Exercise ~ as.factor(Smoke), data = Pulse)
```

