---
title: "BTS 510 Lab 7"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
library(tidyverse)
set.seed(12345)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* **Compare** and **contrast** *point estimates* and *interval estimates*
* **Construct** *confidence intervals*
* **Interpret** *confidence intervals*

## Vary population distribution, sample size, confidence interval level

### Normally distributed: $population \sim \mathcal{N}(\mu = 10, \sigma^2 = 25)$

* Two sample sizes: 30 and 100
* Note that `rnorm()` takes standard deviation as an argument, not variance

```{r}
norm30 <- rnorm(n = 30, mean = 10, sd = 5)
norm100 <- rnorm(n = 100, mean = 10, sd = 5)
```

* Calculate the means, variances, and standard deviations for each sample

```{r}
mean(norm30)
mean(norm100)
var(norm30)
var(norm100)
sd(norm30)
sd(norm100)
```

* Construct the 95% and 99% confidence intervals for each mean

$[\bar{X} - z_{1-\alpha/2}\ \sqrt{\frac{\sigma^2}{n}}, \bar{X} + z_{1-\alpha/2}\ \sqrt{\frac{\sigma^2}{n}}]$

* I'm just having R make the calculations for me, rather than doing by hand
  * I'm also doing it in several steps so you can see where each piece comes from, but you can do it all in one step if you're confident


```{r}
norm30_95moe <- 1.96 * sqrt(var(norm30)/30)
norm30_95lcl <- mean(norm30) - norm30_95moe
norm30_95ucl <- mean(norm30) + norm30_95moe

norm30_99moe <- 2.326 * sqrt(var(norm30)/30)
norm30_99lcl <- mean(norm30) - norm30_99moe
norm30_99ucl <- mean(norm30) + norm30_99moe  

norm100_95moe <- 1.96 * sqrt(var(norm100)/100)
norm100_95lcl <- mean(norm100) - norm100_95moe
norm100_95ucl <- mean(norm100) + norm100_95moe

norm100_99moe <- 2.326 * sqrt(var(norm100)/100)
norm100_99lcl <- mean(norm100) - norm100_99moe
norm100_99ucl <- mean(norm100) + norm100_99moe  
```

* You can look at them in the output, or you can have R print them in your Quarto document * (Look at the code to see how this was done)
  * n = 30, 95%: [`r round(norm30_95lcl, 3)`, `r round(norm30_95ucl, 3)`]
  * n = 30, 99%: [`r round(norm30_99lcl, 3)`, `r round(norm30_99ucl, 3)`]
  * n = 100, 95%: [`r round(norm100_95lcl, 3)`, `r round(norm100_95ucl, 3)`]
  * n = 100, 99%: [`r round(norm100_99lcl, 3)`, `r round(norm100_99ucl, 3)`]

* How do the confidence intervals vary with 1) sample size (i.e., n = 30 vs n = 100) and 2) confidence level (i.e., 95% vs 99%)?


### Bernoulli distributed: $population \sim B(0.3)$

* Two sample sizes: 30 and 100

```{r}
bern30 <- rbinom(30, 1, 0.3)
bern100 <- rbinom(100, 1, 0.3)
```

* Calculate the means, variances, and standard deviations for each sample

```{r}

```

* Construct the 95% and 99% confidence intervals for each mean

$[\bar{X} - z_{1-\alpha/2}\ \sqrt{\frac{\sigma^2}{n}}, \bar{X} + z_{1-\alpha/2}\ \sqrt{\frac{\sigma^2}{n}}]$


```{r}

```

* How do the confidence intervals vary with 1) sample size (i.e., n = 30 vs n = 100) and 2) confidence level (i.e., 95% vs 99%)?

### Binomial distributed: $population \sim Bin(50, 0.3)$

* Two sample sizes: 30 and 100

```{r}
binom30 <- rbinom(30, 50, 0.3)
binom100 <- rbinom(100, 50, 0.3)
```

* Calculate the means, variances, and standard deviations for each sample

```{r}

```

* Construct the 95% and 99% confidence intervals for each mean

$[\bar{X} - z_{1-\alpha/2}\ \sqrt{\frac{\sigma^2}{n}}, \bar{X} + z_{1-\alpha/2}\ \sqrt{\frac{\sigma^2}{n}}]$

```{r}
 
```

* How do the confidence intervals vary with 1) sample size (i.e., n = 30 vs n = 100) and 2) confidence level (i.e., 95% vs 99%)?

### Poisson distributed: $population \sim Poisson(1.5)$

* Two sample sizes: 30 and 100

```{r}
pois30 <- rpois(n = 30, lambda = 1.5)
pois100 <- rpois(n = 30, lambda = 1.5)
```

* Calculate the means, variances, and standard deviations for each sample

```{r}

```

* Construct the 95% and 99% confidence intervals for each mean

$[\bar{X} - z_{1-\alpha/2}\ \sqrt{\frac{\sigma^2}{n}}, \bar{X} + z_{1-\alpha/2}\ \sqrt{\frac{\sigma^2}{n}}]$

```{r}

```

* How do the confidence intervals vary with 1) sample size (i.e., n = 30 vs n = 100) and 2) confidence level (i.e., 95% vs 99%)?

## Summary

* How do the confidence intervals vary with sample size?

* How do the confidence intervals vary with confidence level?

* How do the confidence intervals vary with population distribution?

* Did you notice anything else? 




