---
title: "BTS 510 Lab 11"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
library(tidyverse)
set.seed(12345)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* *Describe* the logic of ANOVA
* *Conduct* and *interpret* the overall ANOVA test
* *Conduct* post hoc tests

## Data

* `Amyloid` dataset (N = 57) in **Stat2Data** package
  * `Group`: mAD=Alzheimer’s, MCI=mild impairment, NCI=no impairment
  * `Abeta`: Amount of Abeta from the posterior cingulate cortex (pmol/g tissue)

```{r}
library(Stat2Data)
data(Amyloid)
head(Amyloid)
```

## Descriptives and data visualization

### Means, variances, and SDs by group

```{r}
Amyloid %>% group_by(Group) %>%
  summarize(mean = mean(Abeta),
            variance = var(Abeta),
            sd = sd(Abeta))
```

* For some reason, these values get rounded to the whole number. You can get the full numbers to print if you use a package to print prettier tables, like **kableExtra** (only works for HTML output) 

```{r}
library(kableExtra)
Amyloid %>% group_by(Group) %>%
  summarize(mean = mean(Abeta),
            variance = var(Abeta),
            sd = sd(Abeta)) %>%
  kable()
```

### Equal variance assumption

* Graphically

```{r}
ggplot(data = Amyloid,
       aes(x = Group, y = Abeta)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(width = 0.2)
```

* Statistical test

```{r}
library(car)
leveneTest(Abeta ~ Group, data = Amyloid)
```

* Variances are roughly equal -- good to go!

## Analysis of variance

* Use both `lm()` with `contr.sum` and `anova()` to get correct type III sums of squares

```{r}
m1 <- lm(Abeta ~ Group,
         contrasts=list(Group='contr.sum'),
         data = Amyloid)
anova(m1)
```

* Overall, there is an effect of `Group`
  * We reject the $H_0$ that all the group means are equal, F(2, 54) = 5.97, p <.01. At least one group mean is different from the others.

### Normality assumption

* Graphically

```{r}
m1resid <- data.frame(residuals(m1))
ggplot(data = m1resid, 
       aes(sample = residuals.m1.)) +
  geom_qq() +
  geom_qq_line()

m1resid <- data.frame(residuals(m1))
ggplot(data = m1resid, 
       aes(x = residuals.m1.)) +
  geom_histogram()
```

* Statistical test

```{r}
shapiro.test(resid(m1))
```

* Normality assumption is violated
  * We should consider non-parametric alternatives that don't assume normality
  * But for the purposes of this class, we're going to carry on anyway

## Post hoc tests

* I'm only going to use the reference / dummy / `treatment` contrasts here
  * Feel free to try out the others

* Dummy coding compares each group to a specific reference group (e.g., control)
  * In R, the default reference group is the **first** group 
  * Factors in R are ordered alphabetically (if character names) or numerically (if number names)
  * So this will do two comparisons (`mAD` is the reference group)
    * `mAD` vs `MCI`
    * `mAD` vs `NCI`

```{r}
m1_trt <- lm(Abeta ~ Group,
         contrasts=list(Group='contr.treatment'),
         data = Amyloid)
summary(m1_trt)
```

* Intercept = mean for reference group
  * Mean for `mAD` = 761.3
* `GroupMCI` is the difference between reference group and `MCI`
  * Mean for `MCI` = 761.3 - 420.2 = 341.1
* `GroupNCI` is the difference between reference group and `NCI`
  * Mean for `NCI` = 761.3 - 425.0 = 336.3
* Compare these values to the group means calculated above
* $t$ tests tell you whether that contrast (difference) is significant
  * Both contrasts are significant here
  * Mean `Abeta` for `mAD` is different from mean `Abeta` for `MCI`
  * Mean `Abeta` for `mAD` is different from mean `Abeta` for `NCI`
  * Report $t$ tests as we've dont before
    * The degrees of freedom for each test is 1

### Multiple comparisons

* If we controlled for multiple comparisons what are our conclusions?
  * Do those differ from the conclusions where we didn't control for multiple comparisons?
  
```{r}
p.adjust(c(0.00354, 0.00392), "holm", 2)
p.adjust(c(0.00354, 0.00392), "BH", 2)
```
  
## Conclusions

In plain language:

1. Are there differences in amyloid beta levels for the different impairment groups?
2. Which groups are different from one another? 