---
title: "BTS 510 Lab 13"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
library(tidyverse)
library(kableExtra)
set.seed(12345)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* *Describe* the logic of factorial ANOVA
  * Including main effects and interactions
* *Conduct* and *interpret* the factorial ANOVA tests
* *Conduct* post hoc tests for the factorial ANOVA

## Data

* `birthwt` dataset in the **MASS** package
  * `race`: mother’s race (1 = white, 2 = black, 3 = other).
  * `smoke`: smoking status during pregnancy.
  * `bwt`: birth weight in grams
* Convert both `race` and `smoke` to factor variables using `mutate()` and `as.factor()`

```{r}
library(MASS)
data(birthwt)
birthwt <- birthwt %>%
  mutate(race = as.factor(race),
         smoke = as.factor(smoke))
head(birthwt)
```

### Research question

* How are mother's race and mother's smoking status related to the birth weight of a baby?

### Means

* Means by `race` and `smoke`

```{r}
bwt_sum <- birthwt %>% 
  group_by(race, smoke) %>%
  summarize(nbwt = length(bwt),
            meanbwt = mean(bwt),
            sdbwt = sd(bwt))
kable(bwt_sum)  
```

* Plot of means

```{r}
ggplot(data = bwt_sum, 
       aes(x = race, 
           y = meanbwt, 
           group = smoke,
           color = smoke)) +
  geom_point() +
  geom_path(aes(group = smoke)) +
  geom_errorbar(aes(ymin = meanbwt - 1.96*sdbwt/sqrt(nbwt), 
                    ymax = meanbwt + 1.96*sdbwt/sqrt(nbwt)), 
                width = 0.2) +
  labs(x = "Race", 
       y = "Birth weight (g)") +
  scale_x_discrete(labels=c("1" = "White", 
                            "2" = "Black",
                            "3" = "Other")) +
  scale_color_discrete(name = "Smoker", 
                       labels = c("No", "Yes")) +
  geom_hline(yintercept = 2500, linetype = "dashed")
```

## ANOVA

### Assumptions: Equal variance

* Do all the groups (6 groups defined by `race` and `smoke`) have **equal variance** on the outcome variable (`bwt`)?

#### Graphically

```{r}
#| code-fold: true
ggplot(data = birthwt,
       aes(x = race, 
           y = bwt, 
           group = interaction(race, smoke))) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(width = 0.2, aes(color = smoke)) +
  geom_hline(yintercept = 2500, linetype = "dashed")
```

#### Levene's Test

* `leveneTest()` in **car** package
  * $H_0$: No differences in variances across groups
  * NS result = equal variances

```{r}
#| code-fold: true
library(car)
leveneTest(bwt ~ race:smoke, data = birthwt)
```

* The assumption of equal variances is met

### ANOVA

* **Step 1**: `lm()` function for **linear model** 
  * `outcome ~ predictors` format
  * Effects of `race`, `smoke`, and interaction (`race*smoke`)
  * Contrasts: Sum (`contr.sum`) in `list()` function to allow one for each factor
* **Step 2**: `Anova()` function in **car** package
  * Provide model (`m1` on last slide)
  * `type = 3` for correct type III sums of squares

```{r}
m1 <- lm(bwt ~ race + smoke + race*smoke,
         contrasts=list(race = contr.sum, 
                        smoke = contr.sum),
         data = birthwt)

library(car)
Anova(m1, type = 3)
```

### Overall conclusions

* We **reject** the null hypothesis that all `race` means are equal, F(2, 183) = 6.18, p = .0025. **At least one `race` group mean is different from the others.**
* We **reject** the null hypothesis that all `smoke` means are equal, F(1, 183) = 7.15, p = .0082. **At least one `smoke` group mean is different from the others.**
* We **retain** the null hypothesis that there is **not** an interaction, F(2, 183) = 2.25, p = .1085. **There is no interaction of `race` and `smoke`.**
  * The effect of `race` doesn't depend on `smoke`
  * The effect of `smoke` doesn't depend on `race`

### Assumptions: Normality

#### Graphically

```{r}
#| code-fold: true
m1resid <- data.frame(residuals(m1))
ggplot(data = m1resid, 
       aes(sample = residuals.m1.)) +
  geom_qq() +
  geom_qq_line()
```

```{r}
#| code-fold: true
m1resid <- data.frame(residuals(m1))
ggplot(data = m1resid, 
       aes(x = residuals.m1.)) +
  geom_histogram()
```

#### Shapiro-Wilk Test

* `shapiro.test()` in **stats**
  * $H_0$: Observations come from normally distributed population
  * NS result = normally distributed residuals

```{r}
#| code-fold: true
shapiro.test(resid(m1))
```

### Post hoc tests

* `emmeans()` function in **emmeans** package
  * Provide model name (`m1`)
  * Type of contrast (`spec = trt.vs.ctrl` for 1st group as reference)
    * Other options: `pairwise`, `trt.vs.ctrlk` (last group as reference)
  * Factor you want contrasts for: here, only `race`
  * Adjustment for multiple comparisons: here, `"none"`
    * Other options: `tukey`, `bonf`, `fdr`
* Note that these are not the **raw means**
  * **Estimated marginal means** (emmeans): Model based

```{r}
library(emmeans)
emmeans(m1, spec = trt.vs.ctrl ~ race, adjust = "none")
```

* What if we used a Bonferroni correction?

```{r}
emmeans(m1, spec = trt.vs.ctrl ~ race, adjust = "bonf")
```

* What about using false discovery rate (FDR)?

```{r}
emmeans(m1, spec = trt.vs.ctrl ~ race, adjust = "fdr")
```

* What if we wanted all possible pairwise comparisons?

```{r}
emmeans(m1, spec = pairwise ~ race, adjust = "none")
```

* What if we wanted all possible pairwise comparisons and also wanted to include `smoke` in this?

```{r}
emmeans(m1, spec = pairwise ~ race*smoke, adjust = "none")
```

* Wow, let's throw a Bonferroni correction on that one...

```{r}
emmeans(m1, spec = pairwise ~ race*smoke, adjust = "bonf")
```

* What if we wanted to use the last group ("other") as the reference?

```{r}
emmeans(m1, spec = trt.vs.ctrlk ~ race, adjust = "none")
```

* Just for fun, let's look at the `smoke` contrast, even though we don't need to because there are only 2 levels
  * Remember that squaring the $t$ statistic here will give you the $F$ statistic from the ANOVA table

```{r}
emmeans(m1, spec = pairwise ~ smoke, adjust = "none")
```

## How to do ANOVA wrong

* There are a LOT of ways to do the ANOVA wrong
  * The most obvious one is to use the `anova()` function
  * This gives type I sums of squares instead of the correct type III

```{r}
anova(m1)
```

* Compare to correct type II SSs

```{r}
library(car)
Anova(m1, type = 3)
```

* What's the same?
  * Degrees of freedom
  * Residual SS: That's what's left over after everything else
  * Sums of squares, etc. for interaction: More on that in a second
* What's different?
  * Everything else

### Why are they different?

* Type I SS *partials out* or holds constant **all effects entered before this effect**
  * **Order you enter effects matters**
  * Effect of race is partialling **nothing**
  * Effect of smoke is partialling **race only**
  * Effect of interaction is partialling **both race and smoke**
* Type III SS *partials out* or holds constant **all other effects**
  * **Order you enter effects doesn't matter**
  * Effect of race is partialling **both smoke and the interaction**
  * Effect of smoke is partialling **both race and the interaction**
  * Effect of interaction is partialling **both race and smoke**

### Change the predictor order

* Change the order of predictors: `smoke` then `race` then interaction

```{r}
m2 <- lm(bwt ~ smoke + race + race*smoke,
         contrasts=list(race = contr.sum, 
                        smoke = contr.sum),
         data = birthwt)
```


* Type I (wrong) model
  * **Different results from original predictor order (`m1`)**

```{r}
anova(m2)
```

* Type III (correct) model
  * **Same results as original predictor order (`m1`)**
  * (Note different order of predictors)

```{r}
library(car)
Anova(m2, type = 3)
```

## How would I actually run this?

* Research question: How are mother's race and mother's smoking status related to the birth weight of a baby?
  * I rarely use an actual ANOVA
  * Linear regression (for continuous outcomes) is more flexible

### Recode grouping variables

* We already recoded the group variables into factors, but I'm going to explicitly create **dummy codes** (also called "indicator variables") for them
  * Each dummy code has a value of 1 for the group it's about
  * e.g., `white` dummy code has a 1 when `race` = 1 and a 0 otherwise
  * e.g., `black` dummy code has a 1 when `race` = 2 and a 0 otherwise
  * e.g., `other` dummy code has a 1 when `race` = 3 and a 0 otherwise
  * I use `mutate()` with `ifelse()` to create these
    * `ifelse()` has 3 arguments:
      * Logical argument (e.g., `race == 1`)
      * What to do if the logical argument is true 
      * What to do otherwise
    * For the new variable `white`:
      * If `race` = 1, then `white` = 1
      * Otherwise, `white` = 0
  * Note that I'm naming each variable after its **group**
  
```{r}
birthwt <- birthwt %>%
  mutate(white = ifelse(race == 1, 1, 0),
         black = ifelse(race == 2, 1, 0),
         other = ifelse(race == 3, 1, 0))
kable(head(birthwt))
```

* `smoke` is already coded 0 and 1, but I'll do the same there
  * I'm calling this variable `smoker`: a value of 1 means they're a smoker

```{r}
birthwt <- birthwt %>%
  mutate(smoker = ifelse(smoke == 1, 1, 0))
kable(head(birthwt))
```

### Use dummy codes as predictors

* You already know everyone's group from any 2 of the 3 dummy codes
  * In general, for $a$ groups, you use $a - 1$ of the variables
  * If you tried to use all three variables, you'd get an error because you have redundant information
    * It's like if you used weight in pounds and weight in kg in the same model
* The dummy code you leave out is the **reference group**
  * Use `black` and `other`: "White" is the reference group
  * Use `white` and `other`: "Black" is the reference group
  * Use `white` and `black`: "Other" is the reference group
  * *You get to just pick your reference group*

### Reference: "white" and "nonsmoker"

```{r}
m1_white <- lm(bwt ~ black + other + smoker + black*smoker + other*smoker,
               data = birthwt)
summary(m1_white)
```

* Intercept: Mean when all predictors = 0 (reference on both)
  * 3428.7 = mean birth weight for white moms (`black` = 0 and `other` = 0) who don't smoke (`smoker` = 0)
  * Significant $t$-test: This value is significantly different from 0 (which is likely not of interest, but could be)
* `black`: Difference between black moms and white moms (`black` = 0 and `other` = 0) for nonsmokers (`smoker` = 0)
  * -574.2 = Difference between white and black moms who don't smoke 
    * 3428.7 - 574.2 = 2854.5 = mean birth weight for black moms who don't smoke
    * Significant $t$-test: These two means are significantly different from one another
* `other`: Difference between other race moms and white moms (`black` = 0 and `other` = 0) for nonsmokers (`smoker` = 0)
  * -613.0 = Difference between white and other race moms who don't smoke 
    * 3428.7 - 613.0 = 2815.7 = mean birth weight for other race moms who don't smoke
    * Significant $t$-test: These two means are significantly different from one another
* `smoker`: Difference between nonsmokers (`smoker` = 0) and smokers (`smoker` = 1) for  white moms (`black` = 0 and `other` = 0)
  * -601.9 = Difference between white smokers and white nonsmokers
    * From above: 3428.7 = mean birth weight for white moms (`black` = 0 and `other` = 0) who don't smoke (`smoker` = 0)
    * 3428.7 - 601.9 = 2826.8 = mean birth weight for white smoker moms
    * Significant $t$-test: These two means are significantly different from one another
* `black:smoker`: *Difference in difference* between white smokers and white nonsmokers vs black smokers and black nonsmokers
  * From above: -601.9 = Difference between white smokers and white nonsmokers 
  * Difference between black smokers and black nonsmokers is 251.4 points *higher* than that, or -601.9 + 251.4 = -350.5
  * Non-significant $t$-test: These two *differences* are **not** significantly different from one another
* `other:smoker`: *Difference in difference* between white smokers and white nonsmokers vs other race smokers and other race nonsmokers
  * From above: -601.9 = Difference between white smokers and white nonsmokers 
  * Difference between other race smokers and other race nonsmokers is 543.3 points *higher* than that, or -601.9 + 543.3 = -58.6
  * Significant $t$-test: These two *differences* are significantly different from one another

#### Intercept and main effects

```{r}
library(grid)
ggplot(data = bwt_sum, 
       aes(x = race, 
           y = meanbwt, 
           group = smoke,
           color = smoke)) +
  geom_point(size = 8) +
  labs(x = "Race", 
       y = "Birth weight (g)") +
  scale_x_discrete(labels=c("1" = "White", 
                            "2" = "Black",
                            "3" = "Other")) +
  scale_color_discrete(name = "Smoker", 
                       labels = c("No", "Yes")) +
  annotate("text", x = 1, y = 3500, label = "3428.7 *") +
  annotate("segment", x = 1, xend = 2, y = 3428, yend = 2854,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
  annotate("text", x = 1.5, y = 3125, label = "-574.2 *") +
  annotate("segment", x = 1, xend = 3, y = 3428, yend = 2815,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
  annotate("text", x = 2.5, y = 3125, label = "-613.0 *") +
  annotate("segment", x = 1, xend = 1, y = 2826, yend = 3428,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
  annotate("text", x = .8, y = 3125, label = "-601.9 *")
```

#### Interaction effects

```{r}
library(grid)
ggplot(data = bwt_sum, 
       aes(x = race, 
           y = meanbwt, 
           group = smoke,
           color = smoke)) +
  geom_point(size = 8) +
  labs(x = "Race", 
       y = "Birth weight (g)") +
  scale_x_discrete(labels=c("1" = "White", 
                            "2" = "Black",
                            "3" = "Other")) +
  scale_color_discrete(name = "Smoker", 
                       labels = c("No", "Yes")) +
  annotate("segment", x = 1, xend = 1, y = 2826, yend = 3428,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
  annotate("text", x = .8, y = 3125, label = "-601.9 *") +
  
  annotate("segment", x = 2, xend = 2, y = 2854, yend = 2504,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
  annotate("text", x = 2, y = 3000, label = "-350.5 = -601.9 + 251.4") +
  
  annotate("segment", x = 3, xend = 3, y = 2815, yend = 2757,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
  annotate("text", x = 3, y = 2900, label = "-58.6 = -601.9 + 543.3") +
  
  annotate("segment", x = 1, xend = 1.95, y = 3500, yend = 3500,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
  annotate("text", x = 1.5, y = 3450, label = "251.4 NS") +
  
  annotate("segment", x = 2.05, xend = 3, y = 3500, yend = 3500,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
  annotate("text", x = 2.5, y = 3450, label = "543.3 *")
```

### Reference: "black" and "nonsmoker"

```{r}
m1_black <- lm(bwt ~ white + other + smoker + white*smoker + other*smoker,
               data = birthwt)
summary(m1_black)
```

* Which means do each of the effects (coefficients) refer to?
  * Use the table of raw means as guide
